---
title: "PSC205A Assignment 01: Matrix Algebra"
format: pdf
---

```{r}
library(lavaan)
library(lavaanPlot)
library(dplyr)
```


```{r}
dat <- read.table("mhs_latent.dat", na.string='.')
names(dat) <- c("id", "female", "age",
                "pc1", "app1", "gw1", "pw1", "par1", "teach1", "mate1", "fri1", "mot1", "aff1", "aut1",
                "pc2", "app2", "gw2", "pw2", "par2", "teach2", "mate2", "fri2", "mot2", "aut2", "aff2", "acomp2", "aca_imp",
                "pc3", "app3", "gw3", "pw3", "par3", "teach3", "mate3", "fri3", "mot3", 'aff3', "aut3", "acomp3",
                "pc4", "app4", "gw4", "pw4", "par4", "teach4", "mate4", "fri4", "mot4", "aff4", "aut4", "acomp4", "roman4")

```

# 1. SEM analysis

```{r}
mod1 <- "
#factor loadings
pc =~ pc1 + pc2 + pc3 + pc4
app =~ app1 + app2 + app3 + app4
gw =~ gw1 + gw2 + gw3 + gw4
pw =~ pw1 + pw2 + pw3
par =~ par1 + par2 + par3 + par4
teach =~ teach1 + teach2 + teach3 + teach4
mate =~ mate1 + mate2 + mate3 + mate4
fri =~ fri1 + fri2 + fri3 + fri4
mot =~ mot1 + mot2 + mot3 + mot4
aff =~ aff1 + aff2 + aff3 + aff4
aut =~ aut1 + aut2 + aut3 + aut4

# structural relations
gw ~ app + pc + teach + par + mate + fri
pw ~ app +pc + teach + par + mate + fri

aff ~ pw
mot ~ pw + aff

# covariances
app ~~ 0*pc
app ~~ 0*teach
app ~~ 0*par
app ~~ 0*mate
app ~~ 0*fri
app ~~ 0*aff
app ~~ 0*mot

pc ~~ 0*teach
pc ~~ 0*par
pc ~~ 0*mate
pc ~~ 0*fri
pc ~~ 0*aff
pc ~~ 0*mot

gw ~~ 0*pw
gw ~~ 0*aff
gw ~~ 0*mot

teach ~~ 0*par
teach ~~ 0*mate
teach ~~ 0*fri
teach ~~ 0*aff
teach ~~ 0*mot

par ~~ 0*mate
par ~~ 0*fri
par ~~ 0*aff
par ~~ 0*mot

mate ~~ 0*fri
mate ~~ 0*aff
mate ~~ 0*mot

fri ~~ 0*aff
fri ~~ 0*mot
"
```

```{r}
fit1 <- cfa(mod1, data = dat, fixed.x=FALSE, missing="fiml")
res1 <- summary(fit1, fit.measures=T, standardized=TRUE)

```

# 2. Additional models

```{r}
mod_ind_m1 <- modindices(fit1)
head(arrange(mod_ind_m1, desc(mi)))
```

The modification indices tell us that the model might be improved by allowing the variables friend support and classmate support to correlate. We can start by freeing this relation.

## 2.1. Model 2: mate ~~ fri

```{r}
mod2 <- "
#factor loadings
pc =~ pc1 + pc2 + pc3 + pc4
app =~ app1 + app2 + app3 + app4
gw =~ gw1 + gw2 + gw3 + gw4
pw =~ pw1 + pw2 + pw3
par =~ par1 + par2 + par3 + par4
teach =~ teach1 + teach2 + teach3 + teach4
mate =~ mate1 + mate2 + mate3 + mate4
fri =~ fri1 + fri2 + fri3 + fri4
mot =~ mot1 + mot2 + mot3 + mot4
aff =~ aff1 + aff2 + aff3 + aff4
aut =~ aut1 + aut2 + aut3 + aut4

# structural relations
gw ~ app + pc + teach + par + mate + fri
pw ~ app +pc + teach + par + mate + fri

pw ~ aff
pw ~ mot

pc ~ mate

# covariances
app ~~ 0*pc
app ~~ 0*teach
app ~~ 0*par
app ~~ 0*mate
app ~~ 0*fri
app ~~ 0*aff
app ~~ 0*mot

pc ~~ 0*teach
pc ~~ 0*par
pc ~~ 0*mate
pc ~~ 0*fri
pc ~~ 0*aff
pc ~~ 0*mot

gw ~~ 0*pw
gw ~~ 0*aff
gw ~~ 0*mot

teach ~~ 0*par
teach ~~ 0*mate
teach ~~ 0*fri
teach ~~ 0*aff
teach ~~ 0*mot

par ~~ 0*mate
par ~~ 0*fri
par ~~ 0*aff
par ~~ 0*mot

mate ~~ fri
mate ~~ 0*aff
mate ~~ 0*mot

fri ~~ 0*aff
fri ~~ 0*mot
"
```

```{r}
fit2 <- cfa(mod2, data = dat, fixed.x=FALSE, missing="fiml")
res2 <- summary(fit2, fit.measures=T, standardized=TRUE)

mod_ind_m2 <- modindices(fit2)
head(arrange(mod_ind_m2, desc(mi)))
```

The new modification indices suggest the model can be further improved by allowing parent support and teacher support to correlate.

## 2.2. Model 3: pc ~ mate + aff

```{r}
mod3 <- "
#factor loadings
pc =~ pc1 + pc2 + pc3 + pc4
app =~ app1 + app2 + app3 + app4
gw =~ gw1 + gw2 + gw3 + gw4
pw =~ pw1 + pw2 + pw3
par =~ par1 + par2 + par3 + par4
teach =~ teach1 + teach2 + teach3 + teach4
mate =~ mate1 + mate2 + mate3 + mate4
fri =~ fri1 + fri2 + fri3 + fri4
mot =~ mot1 + mot2 + mot3 + mot4
aff =~ aff1 + aff2 + aff3 + aff4
aut =~ aut1 + aut2 + aut3 + aut4

# structural relations
gw ~ app + pc + teach + par + mate + fri
pw ~ app +pc + teach + par + mate + fri

pw ~ aff
pw ~ mot

pc ~ mate + aff

# covariances
app ~~ 0*pc
app ~~ 0*teach
app ~~ 0*par
app ~~ 0*mate
app ~~ 0*fri
app ~~ 0*aff
app ~~ 0*mot

pc ~~ 0*teach
pc ~~ 0*par
pc ~~ 0*mate
pc ~~ 0*fri
pc ~~ 0*aff
pc ~~ 0*mot

gw ~~ 0*pw
gw ~~ 0*aff
gw ~~ 0*mot

teach ~~ par
teach ~~ 0*mate
teach ~~ 0*fri
teach ~~ 0*aff
teach ~~ 0*mot

par ~~ 0*mate
par ~~ 0*fri
par ~~ 0*aff
par ~~ 0*mot

mate ~~ fri
mate ~~ 0*aff
mate ~~ 0*mot

fri ~~ 0*aff
fri ~~ 0*mot
"
```

```{r}
fit3 <- cfa(mod3, data = dat, fixed.x=FALSE, missing="fiml")
res3 <- summary(fit3, fit.measures=T, standardized=TRUE)

mod_ind_m3 <- modindices(fit3)
head(arrange(mod_ind_m3, desc(mi)))
```

Now the modification indices suggest an improvement in fit by regressing Physical appearance on Perceived competence.

## 2.3. Model 4: app ~ pc

```{r}
mod4 <- "
#factor loadings
pc =~ pc1 + pc2 + pc3 + pc4
app =~ app1 + app2 + app3 + app4
gw =~ gw1 + gw2 + gw3 + gw4
pw =~ pw1 + pw2 + pw3
par =~ par1 + par2 + par3 + par4
teach =~ teach1 + teach2 + teach3 + teach4
mate =~ mate1 + mate2 + mate3 + mate4
fri =~ fri1 + fri2 + fri3 + fri4
mot =~ mot1 + mot2 + mot3 + mot4
aff =~ aff1 + aff2 + aff3 + aff4
aut =~ aut1 + aut2 + aut3 + aut4

# structural relations
gw ~ pc + teach + par + mate + fri
pw ~ app +pc + teach + par + mate + fri

pw ~ aff
pw ~ mot

pc ~ mate + aff
app ~ pc

# covariances
app ~~ gw
app ~~ 0*pc
app ~~ 0*teach
app ~~ 0*par
app ~~ 0*mate
app ~~ 0*fri
app ~~ 0*aff
app ~~ 0*mot

pc ~~ 0*teach
pc ~~ 0*par
pc ~~ 0*mate
pc ~~ 0*fri
pc ~~ 0*aff
pc ~~ 0*mot

gw ~~ 0*pw
gw ~~ 0*aff
gw ~~ 0*mot

teach ~~ 0*par
teach ~~ 0*mate
teach ~~ 0*fri
teach ~~ 0*aff
teach ~~ 0*mot

par ~~ 0*mate
par ~~ 0*fri
par ~~ 0*aff
par ~~ 0*mot

mate ~~ 0*fri
mate ~~ 0*aff
mate ~~ 0*mot

fri ~~ 0*aff
fri ~~ 0*mot

"
```

```{r}
fit4 <- cfa(mod4, data = dat, fixed.x=FALSE, missing="fiml")
res4 <- summary(fit4, fit.measures=T, standardized=TRUE)

mod_ind_m4 <- modindices(fit4)
head(arrange(mod_ind_m4, desc(mi)))
```

## Model comparison

```{r}
anova(fit1, fit2, fit3, fit4)
```

# 3. Results summary


