---
title: "Lab 7: Logistic Regression & GLM"
subtitle: "PSC 103B"
author: "Marwin Carmo"
format: 
  revealjs:
    scrollable: true
    incremental: true
    code-tools: true
    code-copy: true
    code-line-numbers: true
    code-link: true
    preview-links: true
    slide-number: true
    self-contained: true
    fig-height: 4
    fig-width: 6
    fig-align: center
    margin-left: "0"
    margin-right: "0"
    width: 1400
    callout-icon: false
    # height: 900
    footer: "PSC 103B - Statistical Analysis of Psychological Data"
    logo: "https://github.com/emoriebeck/psc290-data-viz-2022/raw/main/01-week1-intro/02-code/02-images/ucdavis_logo_blue.png"
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
wine <- read.csv("data/wine.csv")
```


## Logistic regression

- Appropriate for binary outcomes, which only takes on values of 0 or 1.

- Oftentimes, 1 corresponds to a "success" of some type.

- Logistic regression predicts the log-odds of success because there is a linear relationship between the log-odds and your predictors.

## Today's dataset

- We are going to demonstrate logistic regression with this dataset on red wine quality, which can be found at <https://archive.ics.uci.edu/ml/datasets/wine+quality>.

- This dataset contains information on characteristics of the wine, such as the acidity, sugar, pH, alcohol content, etc. as well as a rating of the wine as "good" or "bad".

- The rating was created by dichotomizing a quality variable (so that 
anything with a quality score greater than 5 out of 9 was considered good).

## Today's dataset

```{r}
dplyr::glimpse(wine$quality)
```

- We are going to create a second column that assigns this a numerical
value (1 if the quality is good, 0 if it's bad).

- That way, we can be sure our logistic regression is predicting
the log-odds of a wine being good.

:::fragment
```{r}
wine$quality_binary <- ifelse(wine$quality == "good", 1, 0)

dplyr::glimpse(wine$quality_binary)
```
:::

## Simple Logisitic Regression

- Let's fit a logistic regression model with only one predictor -- alcohol content.

- To fit a logistic regression function, we need to use a new function called `glm()`.

- This function, for the most part, is similar to the `lm()`: `outcome var ~ independent var(s)`.

- We just have to specify a family, using `family = "binomial"` (this tells `R` that the outcome variable is binary, and to use logistic regression).

:::fragment
```{r}
simple_logreg <- glm(quality_binary ~ alcohol,
                    data = wine,
                    family = "binomial")
```
:::

---

```{r}
summary(simple_logreg)
```

## Simple Logisitic Regression

$$
\log(Odds_{\text{Quality}}) = -10.76 + 1.06 \times \text{Alcohol}
$$

- The **intercept** of -10.76 means that a red wine that has 0 
alcohol content has an expected log-odds score of -10.76.

- The **slope** means that for every 1-unit increase in alcohol
content, the log-odds of a wine being rated good increase
by 1.06 points.

- But how do we interpret a log-odds?

- It is not intuitive, and that's why we transform the coefficients
to be interpreted in terms of odds ratios. How can we do that?

## Simple Logisitic Regression

$$
Odds_{\text{Good}} = \exp(-10.76 + 1.06 \times \text{Alcohol})
$$

or (using the rules of exponents),

$$
Odds_{\text{Good}} = \exp(-10.76) \times \exp(1.06 \times \text{Alcohol})
$$

## Interpreting the coefficients

- **Intercept**: it is still the expected value when
alcohol is 0. So when a wine has no alcohol content, the expected
odds of being rated good are exp(-10.76), or .00002.

- **Slope**: it is the multiplicative change in the odds for a 1-unit change in Alcohol. So when alcohol content increases by 1, the odds increase
by a factor of (are multiplied by) $e^{1.06}$ = 2.89, or a 289% increase in the odds.

- Because 1.06 is positive, we know that as the alcohol content increases, the probability of a wine being considered good also increase.

## Mean centering the predictors

::: {.notes}
Just like in linear regression, we can mean-center our predictor variables to improve the in-
terpretation of our intercept – because a wine with an alcohol content of 0 is pretty unlikely
:::

```{r}
wine$alcohol_c <- wine$alcohol - mean(wine$alcohol, na.rm = TRUE)
```

```{r}
logreg_centered <- glm(quality_binary ~ alcohol_c,
                      data = wine,
                      family = "binomial")

summary(logreg_centered)
```

## Mean centering the predictors

- Notice that the value of the slope hasn't changed, 
but now our intercept is 0.24 -- what does this mean?

- The expected log-odds of a wine with an average alcohol
content being rated good is 0.24.

- The odds of it being rated good are $e^{0.24}$ = 1.27
so it's more likely to be rated good than it is to be rated bad.

## Predicted probabilities

- What is the probability that the wine will be rated "good"? 

- Let's take the example of a Cabernet Sauvignon, which  generally has an alcohol content of 14%.

:::fragment
```{r}
(logodds <- -10.76 + 1.06*14)
```
:::

- The expected log-odds are 4.08, which again, are hard for us to interpret so let's transform those into odds.

:::fragment
```{r}
(odds <- exp(logodds))
```
:::

## Predicted probabilities

- We can transform these odds into a probability using, $P = \frac{odds}{1 + odds}$

:::fragment
```{r}
odds / (1 + odds)
```
:::

- A wine with 14% alcohol has almost 100% (98.34%) chance of being rated good according to our model.

## Predicted probabilities

- A probability of 0.5 (equal chance) corresponds
to an odds ratio of 1, which corresponds to a log-odds of 0.

- Therefore, if something has a log-odds greater than 0,
then that **increases** your chance of a success (probability
greater than 0.5), and a log-odds less than 0 **decreases**
your chance of success (probability less than 0.5).

## Now you try

:::{.callout-caution title="Exercise"}
Fit a model predicting quality from the amount of chlorides, which can affect how salty a wine tastes. Interpret the coefficients in terms of log-odds and odds.
:::

## Multiple Logistic Regression

- Just like in linear regression, we can add multiple predictors to our logistic regression model.

:::fragment
```{r}
multiple_logreg <- glm(quality_binary ~ alcohol + sulphates,
                      data = wine,
                      family = "binomial")
summary(multiple_logreg)
```
:::

## Multiple Logistic Regression

$$
\log(Odds_{\text{Good}}) = -12.32 + 1.04 \times \text{Alcohol} + 2.67 \times \text{Sulphates}
$$

- **Intercept**: The expected odds of a wine with no alcohol content and no sulphates is exp(-12.32) = .000004.

- **Slope of Alcohol**: Holding the amount of sulphates constant,
increasing the alcohol content by 1 increases the odds of a
wine being rated as good by a factor of exp(1.04) = 2.83.

- **Slope of Sulphates**: Holding the amount of alcohol constant,
increasing the sulphates by 1 unit increases the odds of a
wine being rated as good by a factor of exp(2.67) = 14.44.

## Multiple Logistic Regression with interactions

```{r}
interact_logreg <- glm(quality_binary ~ alcohol + sulphates + I(alcohol*sulphates),
                      data = wine,
                      family = "binomial")
summary(interact_logreg)
```

# General Linear Model

## General Linear Model

- Many of the tests we learned in class can be rewritten as linear regressions with the help of dummy coding.

- Let's revisit our ANOVA example from a few weeks ago
where we wanted to examine bill length differences
between different species of penguins.

- We had to do an F-test to determine whether any of the means
were different, and then a post-hoc test to see which means
were different.

## General Linear Model

- We can fit this model as a regression model with dummy codes
and this might help us make more specific comparisons
right away.

- A dummy code is a variable that assigns a value of 1 
if a person is in one specific group, and 0 otherwise.

- You can manually create these dummy codes, or if you 
fit a regression model with a "factor" variable as a
predictor, R will automatically create dummy codes.

## General Linear Model

```{r}
# Load the palmerpenguins package
library(palmerpenguins)
```

```{r}
dummyreg <- lm(bill_length_mm ~ species, data = penguins)
summary(dummyreg)
```

## General Linear Model

- `R` automatically assigned one group to be our reference group
and that's generally done by choosing the group that comes
first alphabetically (in this case, Adelie).

- This reference group has a score of 0 on the dummy code
variables.

- Now the slopes are the effects of the dummy codes
for the other 2 species.

---

$$
y_i = b_0 + b_1\text{Chinstrap} + b_2\text{Gentoo} 
$$
- If a penguin is from the Adelie species:

$$
y_i = b_0 + b_1(0) + b_2(0) 
$$
- If a penguin is from the Chinstrap species:

$$
y_i = b_0 + b_1(1) + b_2(0) 
$$
- If a penguin is from the Gentoo species:

$$
y_i = b_0 + b_1(0) + b_2(1) 
$$
---
- The intercept now represents the expected bill length
for a penguin that has a 0 on both dummy codes. That is, 38.79 is the average bill length for Adelie
penguins.

- The slopes represent the difference between the mean
of the reference group, and the mean of the group that the 
dummy code is for:
  - Chinstrap penguins have an average bill length 10.04 mm longer than the average bill length of Adelie penguins.
  - Gentoo 
penguins have an average bill length 8.71 mm longer than the average bill length of Adelie penguins.

## General Linear Model: coefficients

- Both these differences are significant! So unlike before where we had to do a post-hoc test to see which groups
were different, we can automatically see which pairs are significantly different.

- However, not all comparisons are represented - we can't say anything about the difference between Chinstrap and Gentoo penguins.