---
title: "Assignment 06"
author: "Marwin Carmo"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(nlme)
library(lmerTest)
library(ggplot2)
```

```{r}
mydata <- read.csv("../data/mydata.csv")
# removing NAs
mydata <- mydata[!is.na(mydata$insomnia_severity), ]
# convert the randomization variable to factor
mydata$randomization <- factor(mydata$randomization)

```

## (1)	Select a variable in your data for modeling over time. (1 variable, at least 3 occasions). Prepare a person-period (i.e., multiple records per person, “long”) data set for use. Use the same variable and data as Assignment 5.

## (2)	Select a time-varying predictor of interest (categorical or continuous)

I will select Anxiety scores as the time-varying predictor.

## (3)	Determine how to best model the time-varying predictor and run the analysis. Consider the following in the modeling process:

### a.	Is time properly scaled?
### b.	How does the predictor affect the DV as specified by your model? Is your model consistent with theory?
### c.	Do you need to center your time-varying predictor? How does centering affect your results?

## (4)	Write out the equations of your final model


$$
\begin{aligned}
\text{Level 1}: Y_{ij} = \; & \pi_{0j} && \text{(Intercept for individual} \ j\text{)}\\
          & + \pi_{1j}Time_{ij} && \text{(Linear slope of} \ Time_{ij} \ \text{for individual} \ j\text{)}\\
          & + \beta_{20}(Time_{ij})^2 && \text{(Fixed quadratic slope of time)}\\
          & + \beta_{30}PM\_Anx_{ij} &&  \text{(Fixed slope of person-mean Anxiety)}\\
          & + \beta_{40}WP\_Anx_{ij} && \text{(Fixed slope of within-person Anxiety)}\\
          & + \epsilon_{ij} \\
\text{Level 2}: \pi_{0j} = \; &\beta_{00} && \text{(Average intercept for the ACT group)}\\
          & + \beta_{01}D_{ACT,j} && \text{(Difference in intercept for the WL group)}\\
          & + \beta_{02}D_{CBT,j} &&\text{(Difference in intercept for the CBT group)}\\
          & + u_{0j} && \text{(Random intercept deviation for individual j)} \\
          \pi_{0j} = \; &\beta_{00} \\
          &+ \beta_{01}D_{G2,j} && \text{(Difference in linear time slope for the WL group)}\\
          &+ \beta_{02}D_{G3,j} && \text{(Difference in linear time slope for the CBT group)}\\
          &+ u_{0j} && \text{(Random linear time slope deviation for individual j)} \\
\end{aligned}
$$

## (5)	Make a table for your final model as would appear in a paper

## (6)	Write a few sentences reporting the results

## (7)	Include the code you used to complete the assignment.

```{r}
## Mean-centering anxiety so now an axiety score of 0 means an average score
mydata$anxiety_c <- c(scale(mydata$anxiety, scale = FALSE))

## Separating the time-varying predictor into person means and within-person deviations
mydata2 <- mydata |> 
  dplyr::with_groups(record_id, dplyr::mutate, PM_anxiety = mean(anxiety)) |> 
  dplyr::mutate(PMC_anxiety = PM_anxiety - mean(PM_anxiety),
                WP_anxiety = anxiety - PM_anxiety)
  
```

```{r}
## Building off from the best model of Assignment 5, I will start adding 
## the fixed effects of PMC and WP anxiety
m1 <- lmer(insomnia_severity ~ 1 + redcap_event_name + I(redcap_event_name^2) +
               redcap_event_name*randomization + PM_anxiety + WP_anxiety +
       (1+ redcap_event_name  |record_id), 
              REML=FALSE, data=mydata2)

## No more random effects can be added to the model due to sample size restrictions
## So I will test models where the time-varying predictor interacts with
## the time-invariant predictor of randomization

m2 <- lmer(insomnia_severity ~ 1 + redcap_event_name + 
             I(redcap_event_name^2) + redcap_event_name*randomization + 
             PM_anxiety*randomization + WP_anxiety*randomization +
             (1+ redcap_event_name|record_id), 
              REML=FALSE, data=mydata2)
anova(m1, m2)

## Since model 2 doesn't fit significantly better than model 1, I will stick
## with the more parsimonious model
```
