---
title: "Assignment 06"
author: "Marwin Carmo"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

```{r}
library(nlme)
library(lmerTest)
library(ggplot2)
```

```{r}
mydata <- read.csv("../data/mydata.csv")
# removing NAs
mydata <- mydata[!is.na(mydata$insomnia_severity), ]
# convert the randomization variable to factor
mydata$randomization <- factor(mydata$randomization)

```

## (1)	Select a variable in your data for modeling over time. (1 variable, at least 3 occasions). Prepare a person-period (i.e., multiple records per person, “long”) data set for use. Use the same variable and data as Assignment 5.

## (2)	Select a time-varying predictor of interest (categorical or continuous)

Anxiety

## (3)	Determine how to best model the time-varying predictor and run the analysis. Consider the following in the modeling process:

### a.	Is time properly scaled?


### b.	How does the predictor affect the DV as specified by your model? Is your model consistent with theory?


### c.	Do you need to center your time-varying predictor? How does centering affect your results?


