---
title: "Real World Data Report No. 1"
author: "YOUR NAME HERE"
date: "Due: 10/27/2024"
output:
  word_document: default
subtitle: "PSC-012Y"
affiliation: The University of California, Davis
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.align = 'center', 
                      fig.width = 5.00, 
                      fig.height = 5.00)
```

# Introduction


## The Research Question(s)

> [Write both of the research questions you selected]

## The Hypotheses

### Research Question 1:

Null Hypothesis
$H_{0}$:

Alternative Hypothesis:
$H_{1}$: 

### Research Question 2:

Null Hypothesis
$H_{0}$:

Alternative Hypothesis:
$H_{1}$: 

## Selecting Your $\alpha$ Level

> For this study, we will be willing to tolerate a $\alpha = ?.??$ which corresponds to a $?\%$ nominal rate of error.

# Methods

## Measures 

> [Should have as many measures as variables you are using in your analyses]
> [3 Measures templates are being provided for you but you may need more or fewer]

### Measure 1.

> [Describe your measure, what is it, is it Categorical? Continuous?]
> [If categorical, is it nominal? Ordinal? How many levels?]
> [If continuous, is it interval? Ratio? What is the range?]

### Measure 2.

> [Describe your measure, what is it, is it Categorical? Continuous?]
> [If categorical, is it nominal? Ordinal? How many levels?]
> [If continuous, is it interval? Ratio? What is the range?]

### Measure 3.

> [Describe your measure, what is it, is it Categorical? Continuous?]
> [If categorical, is it nominal? Ordinal? How many levels?]
> [If continuous, is it interval? Ratio? What is the range?]

## Analytic Plan

# Results

```{r, echo = TRUE}
# library() any packages you need
dat <- read.csv("RW1/music012Y.csv")
```

## Analysis 1 - [Research Question Here]

> Repeat your research question, what are you studying in this section?

```{r, echo = TRUE}
dat$Primary.streaming.service = factor(
  dat$Primary.streaming.service,
  ordered = FALSE
)

streaming <- table(dat$Primary.streaming.service)

chisq.test(streaming)
```

> Add your interpretation here
> Report your $\chi^{2}$ results as:

$\chi^{2}(DF) = ?.???, p = ?.???$

```{r, echo = TRUE}
# Figure Code Here
```

## Analysis 2 - [Research Question Here]

> Repeat your research question, what are you studying in this section?


```{r, echo = TRUE}
# Statistical Analysis Code Here
```

> Add your interpretation here
> Report your $\chi^{2}$ results as:

$\chi^{2}(DF) = ?.???, p = ?.???$

```{r, echo = TRUE}
# Figure Code Here
```

# Discussion

> Discuss your results. What do they mean? What do they tell us about music and depression/anxiety?