---
title: "Lab 5: Post-hoc Tests & Factorial ANOVA"
subtitle: "PSC 103B"
author: "Marwin Carmo"
format: 
  revealjs:
    scrollable: true
    incremental: true
    code-tools: true
    code-copy: true
    code-line-numbers: true
    code-link: true
    preview-links: true
    slide-number: true
    self-contained: true
    fig-height: 4
    fig-width: 6
    fig-align: center
    margin-left: "0"
    margin-right: "0"
    width: 1400
    callout-icon: false
    # height: 900
    footer: "PSC 103B - Statistical Analysis of Psychological Data"
    logo: "https://github.com/emoriebeck/psc290-data-viz-2022/raw/main/01-week1-intro/02-code/02-images/ucdavis_logo_blue.png"
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Today's dataset

- We'll be using a modified version of last week's data set.

- I randomly chose 68 penguins from each species, because the *manual* post-hoc test we are doing requires equal group sizes.

:::fragment
```{r}
penguins_data <- read.csv("https://shorturl.at/gfDYk")
```
:::

## Post-Hoc Tests

- The one-way ANOVA we conducted last week we were interested in whether the different species of penguins had, on average, equal bill lengths, or whether there was some difference in the average bill length among the 3 species.

:::fragment
```{r}
my_anova <- aov(bill_length_mm ~ species, data = penguins_data)
summary(my_anova)
```
:::

- What can we conclude?

## Post-Hoc Tests

- After a significant ANOVA result, we're interested in learning *where* those differences are.

- Post-hoc tests are meant to correct for the many comparisons you are doing, so that your chances of a **Type 1 error** stays at .05.

- There are multiple post-hoc tests available, and they differ in how conservative they are. We'll focus on **Tukey's Honest Significant Difference (HSD) Test**.

## Tukey's HSD

- It computes a value called a **"Critical Difference"**, and so any group differences that are larger than the Critical Difference (CD) in absolute value are significant.

:::fragment
$$
CD = q \times \sqrt{\frac{MS_w}{n}}
$$
:::

- where $n$ is the group size within each group, and $MS_w$ is the Mean Squared Within
you can get from your ANOVA table.

## Find $n$ and $MS_w$

- $n$

:::fragment
```{r}
table(penguins_data$species)

n <- 68
```
:::

- $MS_w$ (tip: look in the ANOVA summary)

:::fragment
```{r}
summary(my_anova)
```

```{r}
MSW <- 9.2
```
:::

---

To find *q*, we can use the function `qtukey()`, that takes 3 main arguments:

  - `p`, which is the probability to the *left* of the critical value,
  - `nmeans`, which is the number of groups we're comparing and,
  - `df`, which is just df within

- If our test is conducted at the .05 level, then the probability *below* the critical value is .95.

::: fragment
:::{.callout-caution title="Exercise"}
Find *q* using the `qtukey()` function. Set the values for the arguments `p`, `nmeans`, and `df`. Save your result to an object called `qvalue`. Tip: you can also find df within in the ANOVA summary
:::
:::

## Find *q*


```{r}
qvalue <- qtukey(p = .95, 
       nmeans = 3, # because we have 3 species
       df = 201) 
qvalue
```

## Tukey's HSD

- Now, we can go ahead and calculate our CD. 

:::fragment
```{r}
cd <- qvalue * sqrt(MSW / n)
cd
```
:::

- How can we get all the groups' means?

:::fragment
```{r}
penguin_means <- tapply(penguins_data$bill_length_mm, 
                       penguins_data$species, 
                       mean, na.rm = TRUE)
penguin_means
```
:::

---

We can subset this vector to get the mean of each group.

:::fragment
```{r}
adelie_mean <- penguin_means[1]
chinstrap_mean <- penguin_means[2]
gentoo_mean <- penguin_means[3]

```

```{r}
adelie_mean - chinstrap_mean
adelie_mean - gentoo_mean
chinstrap_mean - gentoo_mean
```
:::

---

:::{.callout-caution title="Exercise"}
Given our CD of |1.21| which of these differences are significant?
:::

| Comparison | Mean difference |
|------------|:----------:|
|Adelie - Chinstrap | -10.45 |
|Adelie - Gentoo    | -9.22|
|Chinstrap - Gentoo | 1.23 |

- Adelie has a significantly lower average bill length than both Chinstrap and Gentoo.

- Chinstrap has a significantly longer bill length, on average, than Gentoo.

## Tukey's HSD

`R` has a built-in function  for Tukey's HSD, called `TukeyHSD()`, and the only argument you need to give it is the ANOVA object you saved earlier:

```{r}
TukeyHSD(my_anova)
```

# Factorial ANOVA

## Factorial ANOVA

Factorial ANOVA allow us examine in one test: 

- is there an effect of Grouping Variable 1 (at least one mean is different in those groups)? 
- is there an effect of Grouping Variable 2? 

- is there an interaction between the 2 variables (so that the effect of Grouping Variable 1 depends on Grouping Variable 2, or vice versa)?

## Factorial ANOVA

- **Main effect**: When the average outcome changes consistently across different levels of one factor, regardless of the other factors. 
  - Imagine studying how exercise and diet affect weight loss. If people lose more weight with exercise regardless of their diet, that's a main effect of exercise.

## Factorial ANOVA

- **Interaction**: When the effect of one factor changes depending on the level of another factor. 
  - Back to the exercise and diet example, if exercise leads to more weight loss only for people on a specific diet, there's an interaction.

## Factorial ANOVA

- Let's assume we were interested in whether **body mass** differed not only among the 3 penguin **species**, but also whether it differed among **males and females**.

- If we were to do this as 2 separate ANOVAs, we would have 2 sets of hypotheses:

- One for the **species** of penguin:

  - $H_0: \mu_{Adelie} = \mu_{Chinstrap} = \mu_{Gentoo}$.

  - $H_A$: At least one mean is significantly different from the rest.
  
## Factorial ANOVA

- One for the **sex**:

  - $H_0: \mu_{female} = \mu_{male}$

  - $H_A: \mu_{female} \neq \mu_{male}$
  
- And we also have a hypothesis for the **interaction**:

  - $H_0$: There is no interaction between the species of penguin and their sex.

  - $H_A$: There is an interaction between the species of penguin and their sex.
  
## Factorial ANOVA

- In `R` where we "multiply" our two predictors to include an interaction between them.

:::fragment
```{r}
factorial_anova <- aov(body_mass_g ~ species * sex, data = penguins_data)
summary(factorial_anova)
```
:::

- Which results are significant? What does this tell us?

## Factorial ANOVA

### Main effect of species

At least one species has an average body mass that is significantly different from another species.

```{r}
TukeyHSD(factorial_anova)$species
```

## Factorial ANOVA

### Main effect of sex

There is a significant difference between the body mass of male and female penguins.

```{r}
TukeyHSD(factorial_anova)$sex
```

## Factorial ANOVA

### Interaction

- The difference in body mass between male and female penguins depends on the species.

- We can try to  uncover what that interaction is by graphing the means.

- In this graph, we would have one grouping variable on the x-axis and the other grouping variable as different lines on the graph.

- To do this, we will use the function `interaction.plot()`.

## Factorial ANOVA

```{r}
#| fig-align: "center"
#| out-width: 55%

interaction.plot(x.factor = penguins_data$species, # grouping variable on x axis
                 trace.factor = penguins_data$sex, # grouping variable as lines
                 response = penguins_data$body_mass_g, # outcome variable
                 fun = mean, # summary statistic
                 type = "l", # graph lines
                 col = c("black", "red"), xlab = "Species", ylab = "Body Mass (g)", 
                 trace.label = "Sex")
```

## Factorial ANOVA

- Males, on average, weigh more than females.

- Gentoo penguins weigh more than Adelie and Chinstrap, but Adelie and Chinstrap are pretty close to each other.

- The interaction appears to be that the magnitude of the difference is different depending on the species. In particular, males and females have a much smaller difference in the Chinstrap species than the other 2 species.

## Now you try

:::{.callout-caution title="Exercise"}
Run a factorial ANOVA testing the interaction between the penguin `sex` and its `island` in accounting for differences in `body_mass_g`. What can you conclude?
:::

:::fragment
```{r}
factorial_anova_2 <- aov(body_mass_g ~ sex * island, data = penguins_data)
summary(factorial_anova_2)
```
:::

## Now you try

At least penguins from one island have an average body mass that is significantly different from penguins from another island.

```{r}
TukeyHSD(factorial_anova_2)$island
```

Males have significantly higher body mass than females.

```{r}
TukeyHSD(factorial_anova_2)$sex
```