---
title: "Homework 5"
author: "YOUR NAME HERE"
date: "2025-02-13"
output:
  pdf_document: default
subtitle: "PSC-103B"
affiliation: The University of California, Davis
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 10,
                      fig.height = 5,
                      fig.align = 'center')

```

<!-- # Instructions -->

<!-- Please use R/RStudio to complete the following questions. You will submit your filled-out version of this document **as a PDF** on Canvas. Make sure your PDF looks as expected before submitting. Please **include the code you used to generate your answer for each question (when relevant) as well as the final answer and/or relevant output**. -->

<!-- You may consult with your classmates while working on the assignment, but **you must do all the work yourself – everything you turn in must be your own code and words**. Academic dishonesty will not be tolerated.  -->

<!-- Please submit **a pdf version** of this document with your answers on Canvas by **4:00pm on Thursday, March 6th (before lab)**. -->

# Important reminders:

- Do not insert your answer as comments inside a code chunk. They will be cut out when rendered as pdf.
- Instead, enter your answer as plain text after the '>'.
- Check your compiled document before submission to ensure your answers are displayed correctly. 

---

## 1. We’ll be using the observed frequencies that we used in lab of Fall 2022 enrollment across the different colleges of UC Davis. I have repeated this information below:

|College| Observed Frequency|
|:-----|:-----------------:|
|College of Letters & Sciences| 417|
|College of Agricultural and Environmental Sciences|223|
|College of Biological Sciences| 216|
|College of Engineering| 144|
|Total| 1000|

Suppose we were interested in conducting a goodness-of-fit test to determine whether enrollment in these colleges matches the proportions observed in 1993. Here are the frequencies that were observed in 1993. 

|College| Observed Frequency|
|:-----|:-----------------:|
|College of Letters & Sciences| 465|
|College of Agricultural and Environmental Sciences|223|
|College of Biological Sciences| 191|
|College of Engineering| 121|
|Total| 1000|

### What proportion of students were enrolled in each college in 1993? (1 point)

- $P_{CLAS}$ =
- $P_{CAES}$ =
- $P_{CBS}$ = 
- $P_{CoE}$ =

## 2. Use the probabilities of Question 1 to write the null and alternative hypotheses for the Goodness-of-Fit test (2 points). 

> $H_0:$

> $H_1:$

## 3. What are the expected frequencies? (1 point)

- $E_{CLAS}$ = 
- $E_{CAES}$ = 
- $E_{CBS}$ = 
- $E_{CoE}$ = 

## 4. Conduct the chi-square goodness-of-fit test in R. Show your code and output (1 points).

```{r}

```

## 5. Do you reject or fail to reject the null hypothesis? Does 2022 enrollment in the different colleges match the proportions of 1993? (1 points)

> 

## 6. Now suppose we were interested in whether students’ choice of college was related to how long it took them to graduate. The frequencies from the 2015 cohort are provided in a table below.


|Graduated at   | CLAS  |  CA&ES  |  CBS  |  COE  |  Total  |
|:--------------|:-----:|:-------:|:-----:|:-----:|:-------:|
|3 years or less|72     |30       | 28    |5      |135      |
|4 years        |1238   |677      |803    |444    |3162     |
|5 years        |378    |216      |329    |243    |1166     |
|6 years        |51     |43       |38     |482    |614      |
|Total          |1739   |966      |1198   |1174   |5077     |

### What are the null and alternative hypotheses for the test of independence? (2 points)

> $H_0:$

> $H_1:$

## 7. Conduct the chi-square test of independence. Show your code and output (1 points).

```{r}

```

## 8. Do you reject or fail to reject the null hypothesis? What does this lead you to conclude about time to graduation and college? (1 points)

```{r}

```


