---
title: "Homework 2"
author: "YOUR NAME HERE"
date: "2025-01-16"
output:
  pdf_document: default
subtitle: "PSC-103B"
affiliation: The University of California, Davis
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 10,
                      fig.height = 5,
                      fig.align = 'center')

```

<!-- # Instructions -->

<!-- Please use R/RStudio to complete the following questions. You will submit your filled-out version of this document **as a PDF** on Canvas. Make sure your PDF looks as expected before submitting. Please **include the code you used to generate your answer for each question (when relevant) as well as the final answer and/or relevant output**. -->

<!-- You may consult with your classmates while working on the assignment, but **you must do all the work yourself – everything you turn in must be your own code and words**. Academic dishonesty will not be tolerated.  -->

<!-- Please submit **a pdf version** of this document with your answers on Canvas by **2:00pm on Thursday, January 23rd (before lab)**. -->

## Data

For this homework assignment, we will be using a subset of data from the Nerdy Personality
Attributes Scale (NPAS). This is a real scale, and if you’re interested, you can take the quiz
yourself: https://openpsychometrics.org/tests/NPAS/

The original dataset had over 25,000 responses from all over the world. In order to make things
a bit simpler, I chose 200 responses randomly from each continent (in order to prevent the
responses from being too skewed towards one continent, such as the Americas).
There are a bunch of variables in this dataset, but the ones we’ll be interested in for this
assignment are:

- `nerdy_scale`: Participant’s average “nerdiness” score from the NPAS (1-5 scale)
- `TIPI1`: How extraverted or enthusiastic the participant rates themselves (1-7 scale)
- `TIPI3`: How dependable or self-disciplined the participant rates themselves (1-7 scale)
- `nerdy_selfreport`: Do you see yourself as someone who is nerdy? (1-7 scale)


## 1. Please read the NPAS dataset (NPAS.csv) into R using `read.csv()`. Create a dataset that only contains the variables listed above (recall: how do you subset columns from a dataframe?) (.5 pt). Show the first 6 rows (.5 pt) and last 6 rows (.5 pt) of the new dataset. Show your code.

```{r}
# you can load in the data directly using this link
npas <- read.csv("https://shorturl.at/FsHBd")


```


## 2. Show the correlations among all the variables in the new dataset (1pt). Show your code.

```{r}

```

## 3. Test whether the correlation between the participants’ nerdiness scores on the NPAS (`nerdy_scale`) their level of extraversion is significant (1 pt). Report this correlation using APA format (.5 pt). Show your code.

```{r}

```

>


## 4. Create a scatterplot to reveal the relationship between participants’ level of self-discipline and their nerdiness score on the NPAS (make sure to display their level of self-discipline on the x-axis and nerdiness score on the y-axis, and include appropriate axis labels and a title) (1 pt). Use a few words to describe the general trend revealed from the plot (1 pt). Show your code.

*Note*: This plot might look a little weird (e.g., a bunch of straight lines), and that is because the
level of self-discipline is an ordinal variable, so it cannot take on all possible values between 1
and 7, only the integers.

```{r}

```

>

## 5. Build up a linear model to test if level of self-discipline can significantly predict the nerdiness score (1 pt). Show your code and the summary of the model.

```{r}

```

## 6. Interpret the intercept (.5 pt) and slope (.5 pt) of your model.

>

>

## 7. Report and interpret the (unadjusted) $R^2$ value of your model (1 pt).


## 8. Report the results of this linear regression in APA format (.5 pt).

>

## 9. Assuming a person gives themselves a self-discipline rating of 4, what is their expected nerdiness score on the NPAS based on the linear model from Q5? (.5 pt).