---
title: "PSC 103B - Lab 1 Assignment"
subtitle: "Answer Key"
format: pdf
filters:
  - color.lua
---

# Data

For this homework assignment, we will be using a subset of data from the Nerdy Personality
Attributes Scale (NPAS). This is a real scale, and if you’re interested, you can take the quiz
yourself: https://openpsychometrics.org/tests/NPAS/

The original dataset had over 25,000 responses from all over the world. In order to make things
a bit simpler, I chose 200 responses randomly from each continent (in order to prevent the
responses from being too skewed towards one continent, such as the Americas).
There are a bunch of variables in this dataset, but the ones we’ll be interested in for this
assignment are:

- `nerdy_scale`: Participant’s average “nerdiness” score from the NPAS (1-5 scale)
- `TIPI1`: How extraverted or enthusiastic the participant rates themselves (1-7 scale)
- `TIPI3`: How dependable or self-disciplined the participant rates themselves (1-7 scale)
- `nerdy_selfreport`: Do you see yourself as someone who is nerdy? (1-7 scale)

# Question 1

Please read the NPAS dataset (NPAS.csv) into R using read.csv(). Create a dataset that only contains the variables listed above (recall: how do you subset columns from a dataframe?) (.5 pt). Show the first 6 rows (.5 pt) and last 6 rows (.5 pt) of the new dataset. Show your code.

```{r, echo=TRUE}

npas <- read.csv("data/NPAS.csv")
npas_subset = npas[ , c("nerdy_scale", "TIPI1", "TIPI3", "nerdy_selfreport")]

head(npas_subset, n=6)
tail(npas_subset, n=6)

```
# Question 2

Show the correlations among all the variables in the new dataset (1pt). Show your code.

```{r, echo=TRUE}
round(cor(npas_subset, use = "complete.obs"), 3)
# Note: the round() function is not necessary, but I used it to make my output prettier
```

# Question 3

Test whether the correlation between the participants’ nerdiness scores on the NPAS (`nerdy_scale`) their level of extraversion is significant (1 pt). Report this correlation using APA format (.5 pt). Show your code.

```{r, echo=TRUE}
cor.test(npas_subset$nerdy_scale, npas_subset$TIPI1, 
         method = "pearson", alternative = "two.sided")
```

[We examined the relation between extraversion and participants’ nerdiness score on the NPAS
in a sample of 1000 participants. The Pearson's product-moment correlation between extraversion and nerdiness is
negative, statistically significant, and small, *r* = --0.26, *t*(998) = --8.58, *p* < .001.]{color=#b22222}

# Question 4

Create a scatterplot to reveal the relationship between participants’ level of self-discipline and
their nerdiness score on the NPAS (make sure to display their level of self-discipline on the x-
axis and nerdiness score on the y-axis, and include appropriate axis labels and a title) (1 pt). Use
a few words to describe the general trend revealed from the plot (1 pt). Show your code.

*Note*: This plot might look a little weird (e.g., a bunch of straight lines), and that is because the
level of self-discipline is an ordinal variable, so it cannot take on all possible values between 1
and 7, only the integers.

```{r, echo=TRUE}
plot(npas_subset$TIPI3, npas_subset$nerdy_scale, xlab = "Self-Discipline Rating", 
     ylab = "Nerdiness Score", 
     main = "Relation Between Self-Discipline \nand Nerdiness")
```

[The plot suggests no correlation exists between nerdiness scores and self-discipline ratings, regardless of their level.]{color=#b22222}

# Question 5

Build up a linear model to test if level of self-discipline can significantly predict the nerdiness
score (1 pt). Show your code and the summary of the model.

```{r echo=TRUE}
nerdy_model <- lm(nerdy_scale ~ TIPI3, data = npas_subset)
summary(nerdy_model)
```

# Question 6

Interpret the intercept (.5 pt) and slope (.5 pt) of your model.

[The intercept is 3.67, which means that when a person obtains a score of 0 on the self-
discipline rating, their expected nerdiness score is 3.67.]{color=#b22222}

[The slope is 0.007, which means that for every 1-point increase on the rating of self-discipline, a
participant’s nerdiness score is expected to increase by 0.007.]{color=#b22222}

# Question 7

Report and interpret the (unadjusted) $R^2$ value of your model (1 pt).

[The $R^2$ value is 0.0004, which means that approximately 0.04% of the variation in nerdiness
scores is explained by participants’ self-discipline scores.]{color=#b22222}

# Question 8

Report the results of this linear regression in APA format (.5 pt).

[We ran a simple linear regression to determine whether self-discipline rating predicts nerdiness score on the NPAS. Self-discipline ratings did not significantly predict nerdiness scores (*b* = 0.007, 95% CI
[-0.01, 0.03], *t*(998) = 0.64, *p* = .523). Self-discipline explains a statistically not significant and very weak proportion (0.04%) of variance in nerdiness scores.]{color=#b22222}

# Question 9

Assuming a person gives themselves a self-discipline rating of 4, what is their expected nerdiness score on the NPAS based on the linear model from Q5? (.5 pt).

$$
\begin{aligned}
\widehat{Nerdiness_i} &= b_0 + b_1 \times SelfDiscipline_i\\
\widehat{Nerdiness_i} &= 3.67 + 0.007 \times 4 \\
\widehat{Nerdiness_i} &= 3.70
\end{aligned}
$$

[Their expected nerdiness score would be 3.70]{color=#b22222}