---
title: "PSC 103B - Lab 1 Assignment"
subtitle: "Answer Key"
format: pdf
filters:
  - color.lua
---

# Question 1

Please read the NPAS dataset (NPAS.csv) into R using read.csv(). Create a dataset that only contains the variables listed above (recall: how do you subset columns from a dataframe?) (.5 pt). Show the first 6 rows (.5 pt) and last 6 rows (.5 pt) of the new dataset. Show your code.

```{r, echo=TRUE}
npas <- read.csv("data/NPAS.csv")
npas_subset = npas[ , c("nerdy_scale", "TIPI1", "TIPI3", "nerdy_selfreport")]
head(npas_subset)
tail(npas_subset)

```
# Question 2

Show the correlations among all the variables in the new dataset (1pt). Show your code.

```{r, echo=TRUE}
round(cor(npas_subset, use = "complete.obs"), 3)
```

# Question 3

Test whether the correlation between the participants’ nerdiness scores on the NPAS (nerdy_scale) their level of extraversion is significant (1 pt). Report this correlation using APA format (.5 pt). Show your code.

```{r}
cor_estimate <- cor(npas_subset$nerdy_scale, npas_subset$TIPI1)

sample_size <- nrow(npas_subset)

cor_se <- sqrt((1 - cor_estimate^2) / (sample_size - 2))

t_stat <- cor_estimate / cor_se

df_cor <- sample_size -2

# calculate the p value

pt(t_stat, df = df_cor, lower.tail = FALSE) * 2
```
```{r}
cor.test(npas_subset$nerdy_scale, npas_subset$TIPI1, method = "pearson") -> ct
```

lower.tail = FALSE -> P[X > x]
lower.tail = TRUE -> P[X <= x]


```{r}
cor_estimate <- cor(npas_subset$nerdy_scale, npas_subset$TIPI3)

sample_size <- nrow(npas_subset)

cor_se <- sqrt((1 - cor_estimate^2) / (sample_size - 2))

t_stat <- cor_estimate / cor_se

df_cor <- sample_size -2

# calculate the p value

pt(t_stat, df = df_cor, lower.tail = FALSE) * 2
```
```{r}
cor.test(npas_subset$nerdy_scale, npas_subset$TIPI3, method = "pearson", alternative = "two.sided")
```

