---
title: "PSC 103B Homework 7"
subtitle: "Answer Key"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
wine <- read.csv("../lab/data/wine.csv")
```

## Question 1

```{r}
wine$quality_binary = ifelse(wine$quality == "good", 1, 0)

mod1 = glm(quality_binary ~ residual.sugar, data = wine, family = "binomial")
summary(mod1)

```

## Question 2

$$
\log(Odds) = 0.15 - 0.003 \times \text{ResidualSugar}
$$

The intercept of 0.15 means that a wine with 0 residual sugar has a log-odds of 0.15 of being rated good.

The slope of -0.003 means that for every 1-unit increase in the residual sugar, the log-odds of the wine being rated good decrease by 0.003

## Question 3

$$
\text{Odds} = e^{0.15 - 0.003 \times \text{ResidualSugar}}
$$
The intercept is 1.16, so when a wine has 0 residual sugar, the odds of it being rated good are 1.16 (it is 1.16 times as likely to be rated good as it is to be rated bad).

The slope is 0.997, so for every 1-unit increase in residual sugar, the odds of the wine being rated good change by a factor of 0.997.

## Question 4

```{r}
wine$residual.sugar_c = wine$residual.sugar - mean(wine$residual.sugar)
mod2 = glm(quality_binary ~ residual.sugar_c, data = wine, family = "binomial")
summary(mod2)

```

## Question 5

The intercept in the log-odds is 0.14, which means that a wine with an average amount of residual sugar has a log-odds of 0.14 of being rated good. 

The intercept in terms of the odds is 1.15, which means that a wine with an average amount of residual sugar is 1.15 times as likely to be rated good as it is to be rated bad.

## Question 6

- $\log(Odds) = 0.15 - 0.003 \times 13 = 11$

- $\text{Odds} = e^{0.11} = 1.12$

- $P = \frac{1.12}{1+1.12} = 0.53$

A wine with a residual sugar content of 13 has a log-odds of 0.11 of being rated good, is 1.12 times as likely to be rated good as it is to be rated bad, and has a 53% probability of being rated good. 

## Question 7

```{r}
mod3 = glm(quality_binary ~ residual.sugar + citric.acid, data = wine, family = binomial)
summary(mod3)

```

## Question 8

A wine with 0 residual sugar and 0 citric acid has a log-odds of -0.23 of being rated good.

Every 1-unit increase in residual sugar, holding citric acid constant, decreases the log-odds by 0.04.

Every 1-unit increase in citric acid, holding residual sugar constant, increases the log-odds by 1.71.

## Question 9

A wine with 0 residual sugar and 0 citric acid is 0.79 times as likely to be rated good as it is to be rated bad.

For every 1-unit increase in residual sugar, holding citric acid constant, the odds of a wine being rated good change by a factor of 0.96.

For every 1-unit increase in citric acid, holding residual sugar constant, the odds of a wine being rated good change by a factor of 5.53.

## Question 10

```{r}
library(palmerpenguins)
```



```{r}
dummymodel = lm(bill_length_mm ~ island, data = penguins)
summary(dummymodel)

```
## Question 11

The intercept is 45.26, which represents the average bill length of the reference group (Biscoe Island penguins).

The slope of -1.09 means that Dream Island penguins have an average bill length that is 1.09 mm shorter than the average bill length of Biscoe Island penguins. But this difference is not significant.

The slope of -6.31 means that Torgersen Island penguins have an average bill length that is 6.31 mm shorter than the average bill length of Biscoe Island penguins. This difference is significant.



