---
title: "Homework Week 6"
author: "name"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
---

```{r, message = FALSE, echo = FALSE}
## Load packages
library( bayesplot )
library( rstan )
library(brms)
# use multiple cores
options(mc.cores =  parallel::detectCores())
# compile model and save it 
rstan_options(auto_write = TRUE)
```

# Question 1

Code up the model from Week 4 in stan where our model ($\mathcal{M}1$) was: 

\begin{align}
y \sim & N(\mu, \sigma) \quad [\text{Likelihood}] \\
 & \mu ~ \sim N(\mu_0 = 100, \sigma_0 = 10) \quad [\text{Prior}]\\ 
 & \sigma = 15   \quad [\text{Constant}]
\end{align}

There's two ways to do this in a rmarkdown:

   1) You can write the model into an rmarkdown chunk that calls `stan` and provides the `output.var=` as the model name:  "model1". This calls `stan_model()` for you in the background.
   2) Write a 'model1.stan' file and load and compile it with 'stan_model(file = "model1.stan" )'

The model should also contain a `generated_quantities{}` block that generates values for the predictive distribution 'y_rep' and computes the log likelihood for each value as 'log_lik'. 

In your code, remove the 'eval = FALSE' or set it to 'TRUE', otherwise the chunk will not be evaluated.
```{stan output.var = "model1", eval=TRUE}

// Paste the model1.stan code here:
data {
  int<lower=1> N;
  vector[N] y;
  real<lower=0> sigma;
}

parameters {
  real mu;
  
}

model {
  // Prior
  mu ~ normal(100, 10);
  // likelihood
  y ~ normal(mu, sigma);
}

generated quantities {
  real y_rep[N];
  real log_lik[N];
  for( i in 1:N ){
	y_rep[i] = normal_rng( mu, sigma );
  }
  for(n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | mu, sigma);
  }
}
```

If you save the code in an external file 'model1.stan', do this instead:
```{r eval=FALSE}
## Load and compile the model
model1 <- stan_model(file = "./model1.stan" )
```

# Question 2
Now sample from the  model given this data:
```{r}
set.seed(433)
y <- rbinom(30, 200, .65)
set.seed(NULL )
```

Save the data as a list, so that stan can sample your model: 
```{r}
## Create a data list:
stan_data <- list(y = y,
                  N = length(y),
                  sigma = 15)
## Sample:
fitted <- sampling(model1, data = stan_data )
```

Print the model results, parse the table so that only "mu" is printed:
```{r}
## Results:
summary(fitted, pars = "mu")$summary
```

# Question 3
Now, adapt your stan code to represent this model:

\begin{align}
y \sim & N(\mu, \sigma) \quad [\text{Likelihood}] \\
 & \mu ~ \sim N(\mu_0 = 100, \sigma_0 = 10) \quad [\text{Prior on }\mu]\\ 
 & \sigma ~ \sim \Gamma(10, 1 )   \quad [\text{Prior on }\sigma]
\end{align}

and save the file into "model2.stan"

```{stan output.var = "model2", eval = TRUE}
// Stan model 2 code:
data {
  int<lower=1> N;
  vector[N] y;
}

parameters {
  real mu;
  real<lower=0> sigma;
}

model {
  // Prior
  mu ~ normal(100, 10);
  // likelihood
  y ~ normal(mu, sigma);
}

generated quantities {
  real y_rep[N];
  real log_lik[N];
  for( i in 1:N ){
	y_rep[i] = normal_rng( mu, sigma );
  }
  for(n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | mu, sigma);
  }
}
```

Sample and print output:
```{r}
## Code:

stan_data2 <- list(y = y,
                  N = length(y))

fitted2 <- sampling(model2, data = stan_data2 )

kableExtra::kbl(summary(fitted2)$summary)
```

# Question 4
Now extract both posterior samples with `ppd_mod1 <- extract(fit1, "y_rep" )` which creates a list. The quantities of interest are in `ppd_mod1$y_rep`. Plot the posterior predictive distribution of both models (model1 and model2). Check out the function 'ppc_dens_overlay()' from the 'bayesplot' package. 

```{r}
## Code:
## Draws from the posterior predictive distribution 
ppd_mod1 <- extract(fitted, "y_rep" )
ppd_mod2 <- extract(fitted2, "y_rep" )

color_scheme_set("brightblue")
ppc_dens_overlay(y, ppd_mod1$y_rep)

ppc_dens_overlay(y, ppd_mod2$y_rep)
```

# Question 5

Write the same model, as in Question 3, with brms and recreate the plot for that model as in Question 4:

```{r}
## Code:

brm_mod <- brms::brm(y ~ 1,
          family = gaussian(),
          data = as.data.frame(stan_data),
          prior = prior(normal(100, 10), class = Intercept) +
            prior(gamma(10, 1), class = sigma),
          file = "model_hw6.rds")


```


```{r}

brm_mod <- readRDS("model_hw6.rds")

pp <- brms::posterior_predict(brm_mod)

ppc_dens_overlay(y, pp)

```

