---
title: "Assignment 02"
author: "Your name"
date: "2025-04-16"
output: officedown::rdocx_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


```{r echo=TRUE}
library(ggplot2)
library(kableExtra)
library(psych)
library(moments)
```


```{r echo=TRUE}
mydata <- read.csv("../data/mydata.csv")
```


```{r echo=TRUE}
sel_data <- mydata[, c("record_id", "redcap_event_name", "insomina_severity")]


mydata2_wide <- reshape(sel_data,
                   timevar = "redcap_event_name",
                   idvar = "record_id",
                   direction = "wide")

mydata2 <- mydata2_wide[complete.cases(mydata2_wide), ]
```

# Residualized Gain Scores

## Write out the equation for the residualized gain model.


\[
\begin{aligned}
y_{2i} &= \beta_0 + \beta_1 y_{1i} + e_i \\
y_{3i} &= \beta_0 + \beta_1 y_{2i} + e_i
\end{aligned}
\]


## Run the model for pairs of consecutive occasions (T1 and T2; T2 and T3) and save the residualized gain scores

```{r}
## regressing time 2 on time 1

m1 <- lm(insomina_severity.2 ~ insomina_severity.1, data=mydata2) 
# saving the residuals
res_m1 <- resid(m1)

## regressing time 3 on time 2

m2 <- lm(insomina_severity.3 ~ insomina_severity.2, data=mydata2) 
# saving the residuals
res_m2 <- resid(m2)
```

## Create a table summarizing results of the models

```{r echo=FALSE}
modelsummary::modelsummary(m1,
                           #fmt = 1,
                           shape = term ~ statistic,
                           estimate  = c(
                                         "{estimate}{stars}"),
                           statistic = c("conf.int",
                           "{std.error}", 
                           "{statistic}",
                           "{p.value}"),
                           gof_map = "r.squared")
```

```{r echo=FALSE}
modelsummary::modelsummary(m2,
                           #fmt = 1,
                           shape = term ~ statistic,
                           estimate  = c(
                                         "{estimate}{stars}"),
                           statistic = c("conf.int",
                           "{std.error}", 
                           "{statistic}",
                           "{p.value}"),
                           gof_map = "r.squared")
```

## Summary Statistics for the residualized gain scores

```{r echo=FALSE}

psych::describe(res_m1) |> 
  as.data.frame() |> 
  kable(digits = 1,
        caption = "Residualized gain scores for T2 - T1")

psych::describe(res_m2) |> 
  as.data.frame() |> 
  kable(digits = 1,
        caption = "Residualized gain scores for T3 - T2")
```
# Trend Scores

## Equations

```{r }
poly(c(0, 1.5, 6), degree = 2)
```

\[
\begin{aligned}
T_{\text{linear}} &= (-0.566)Y_1 + (-0.226)Y_2 + (0.793)Y_3\\
T_{\text{quadratic}} &= (0.588)Y_1 + (-0.784)Y_2 + (0.196)Y_3
\end{aligned}
\]

## Compute the linear and quadratic trend scores for your data. 

```{r echo=TRUE}

# Coefficients for linear and quadratic trends
linear <- c(-0.679, -0.226, 0.793)
quadratic <- c(0.588, -0.784, 0.196)


mydata2$linear_trend <- with(mydata2,
  insomina_severity.1 * linear[1] +
  insomina_severity.2 * linear[2] +
  insomina_severity.3 * linear[3]
)

mydata2$quadratic_trend <- with(mydata2,
  insomina_severity.1 * quadratic[1] +
  insomina_severity.2 * quadratic[2] +
  insomina_severity.3 * quadratic[3]
)

```

```{r}
summary_data <- mydata2[, c("insomina_severity.1", 
                            "insomina_severity.2", 
                            "insomina_severity.3", 
                            "linear_trend", 
                            "quadratic_trend")]
colnames(summary_data) <- c("Time1", "Time2", "Time3", "Linear", "Quadratic")

psych::describe(summary_data) |> 
  as.data.frame() |> 
  kable(digits=1)
```

```{r}
round(cor(summary_data, use = "pairwise.complete.obs"), 2) |> 
  as.data.frame() |> 
  kable()
```

# Individual Curves

## Write out the equation for obtaining the individual curve.

\[
y_{it} = \beta_{0i} + \beta_{1i}t + e_{it}
\]

## Compute the individual curves parameters (Intercept and Slope) for your data 

```{r echo=TRUE}

mydata2_long <- reshape(mydata2,
                        varying = list(c("insomina_severity.1", "insomina_severity.2", "insomina_severity.3")),
                        v.names = "insomina_severity",
                        timevar = "time",
                        times = c(1, 2, 3),
                        idvar = "record_id",
                        direction = "long")


istats <- data.frame(id=unique(mydata2_long$record_id),
                     iintercept=rep(NA,length(unique(mydata2_long$record_id))),
                     islope=rep(NA,length(unique(mydata2_long$record_id))))

# run individual regressions
for (i in unique(mydata2_long$record_id)){
    pos <- which(istats$id == i)
    datai <- mydata2_long[which(mydata2_long$record_id == i),] # subset data for each individual

    istats$iintercept[pos] <- coefficients(lm(insomina_severity~time,data=datai))[1]
    istats$islope[pos] <- coefficients(lm(insomina_severity~time,data=datai))[2]
    rm(datai)
}
```

```{r}
means <- sapply(istats[, c("iintercept", "islope")], mean)
sds   <- sapply(istats[, c("iintercept", "islope")], sd)

cor_is <- cor(istats$iintercept, istats$islope)

summary_table <- data.frame(
  Parameter = c("Intercept", "Slope"),
  Mean = round(means, 2),
  SD = round(sds, 2)
)

print(summary_table)
cat("\nCorrelation between intercept and slope:", round(cor_is, 2), "\n")

```

## Plot for predicted curves

```{r}
# merge istats with verb_long
mydata2_long <- merge(mydata2_long,istats,by.x="record_id", by.y = "id")

mydata2_long$ipred <- mydata2_long$iintercept+mydata2_long$islope*mydata2_long$time + 0

# plot predicted individual change
p <- ggplot(data = mydata2_long, aes(x = time, y = ipred, group = record_id))
p + geom_point(color="blue") + geom_line(color="blue") +theme_minimal()
```

## Write a few sentences reporting the results and their interpretation

On average, participants had an intercept of 21.68 (SD = 4.96), indicating a relatively high baseline level of insomnia severity. The mean slope was –3.54 (SD = 2.69), suggesting that most individuals experienced a decrease in insomnia severity over time. The correlation between intercept and slope was –0.61, indicating that participants with higher initial insomnia severity tended to show greater improvements (i.e., more negative slopes). 