---
title: "Assignment01"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
library(ggplot2)
library(kableExtra)
```


```{r}
mydata <- read.csv("../data/mydata.csv")
```

# Question 1

The data set comprises longitudinal data from a sample of 227 participants suffering from insomnia complaints who completed repeated assessments of insomnia severity, anxiety, and depression at three distinct occasions: baseline, 6 weeks (post-treatment), and 6 months (follow-up). Each participant was randomized to one of two therapy groups or the control group: (a) ACT-I; *n* = 76, (b) CBT-I; *n* = 76, or (c) Wait List; *n* = 75. In total, there were 173 (76%) female participants and the mean age was 40.6 years (SD = 10.2). 

# Question 2  

For the time varying variables we can look at Insomnia Severity (ISI). Higher scores mean more insomnia symptoms.

# Question 3

```{r}
sel_data <- mydata[, c("record_id", "redcap_event_name", "insomina_severity")]


mydata2_wide <- reshape(sel_data,
                   timevar = "redcap_event_name",
                   idvar = "record_id",
                   direction = "wide")

psych::describe(mydata2_wide[, c(2,3,4)]) |> 
  kable(digits=2)
```

```{r}
cor(mydata2_wide[, c(2,3,4)], use = "complete.obs") |> 
  kable(digits=2)
```


```{r, fig.dpi=300, fig.height=4, fig.width=6, echo=TRUE}
mydata |> 
  tidyr::pivot_longer(cols = insomina_severity:anxiety,
                      names_to = "symptom", values_to = "score") |> 
  dplyr::filter(symptom == "insomina_severity") |> 
  ggplot(aes( x = redcap_event_name, y = score, group = record_id)) +
  geom_point() + 
  geom_line() +
  labs(x = "Time", y = "Insomnia Severity") +
  scale_x_continuous(breaks = c(1,2,3)) +
  theme_minimal()
```


