---
title: "PSC 204B - Winter 2024 - Homework 6"
author: '"Your Name"'
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
subtitle: Multilevel Modeling
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(foreign)
library(haven)
library(performance)
library(lme4)

```

We will continue to use the *popularity* dataset we used in lab, as well as a new dataset for a repeated measures example. The second dataset comes from Chapter 10 of *Longitudinal Analysis* by Lesa Hoffman, and contains data from 207 older adults without a dementia diagnosis who were assessed roughly every two years for eight years on their memory (*prose recall*). 

```{r}
popular = read_spss("popular2.sav")

memory = read_sas("SAS_Chapter10a.sas7bdat")
```

# Question 1

Use the *popularity* dataset to answer the following questions. In these questions, we are interested in predicting **teacher-rated** popularity of students (*popteach*) based on their gender. 

## Part a) [1 point]

Calculate the intra-class correlation coefficient for how much variability in teacher-rated popularity there is across classrooms. Report the value, and state whether this is considered a large amount of variation (provide a reason, based on cut-offs provided in lecture). 
```{r}
# code
```

**ICC Value:**
**Is this substantial?**: 

## Part b) [1.5 point]

Run 3 regression models, each predicting teacher-rated popularity using gender, and report the output:

- A OLS regression model
- A multilevel model with a random intercept only
- A multilevel model with a random intercept and random slope

```{r}
# code
```

## Part c) [1 point]

How do the 3 models differ in terms of what is / is not allowed to differ across classrooms? In other words, what does it mean to include no random effects, a random intercept only, or a random intercept and random slope?



## Part d) [1 point]

Report and interpret the fixed effects from the model with the random intercept and random slope. 

**(Intercept)**: 

**Effect of sex**: 

## Part e) [1 point]

How do the fixed effects and standard errors differ between the OLS model and the model with a random intercept and random slope?

**Fixed effect of intercept:**

**SE of intercept:**

**Fixed effect of gender:**

**SE of gender:**

# Question 2

## Part a) [1 point]

Create a plot to examine how prose recall (*recall*) changes with age (*tvage*), with age on the x-axis and prose recall on the y-axis. Be sure that observations from the same individual are connected with a line. Be sure to label your axes appropriately. 

```{r}
# code here
```


## Part b) [2 point]

Run a multilevel model to predict prose recall from age, including a random intercept and random slope. Be sure to include person-centered age (centered at age at the initial timepoint) and age at the initial timepoint as predictors. Note that age at the initial timepoint is already provided in the dataset (*ageT0*). 

```{r}
# code here
```


## Part c) [1.5 points]

Report and interpret the fixed effects of the model. 

**(Intercept):**

**Fixed effect of age at the initial timepoint:**

**Fixed effect of age:**


# Extra Credit

## EC 1 [0.5 point]

Using the ICC calculated in Question 1a, calculate the cluster effect. (*Note: each class has 100 students in it*)

```{r}
# code
```


## EC 2 [0.5 point]

Using the model run in Question 1b, with a random intercept and random slope - report and interpret the correlation between random effects.





