---
title: "Lab 07 - Word Clouds and Networks"
author: "YOUR NAME HERE"
date: "11/21/2024"
output:
  html_document:
    df_print: paged
subtitle: "PSC-012Y"
affiliation: The University of California, Davis
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
packet.list = c("tidyverse", "tidytext", "wordcloud", "qgraph",
                "RColorBrewer", "plyr", "devtools", "igraph",
                "GGally", "igraphdata")
# Install Packages
invisible(sapply(packet.list, function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
  }
  library(package, character.only = TRUE)
}))
devtools::install_github("bradleyboehmke/harrypotter")
devtools::install_github("lchiffon/wordcloud2")
library(harrypotter)
library(wordcloud2)
```


# Part 1. Coding and Visualization

## 1. Create a `list()` called `book` and assign it **all** of the Harry Potter books as in this week's Virtual Tutorial [1pts]

```{r}


```

## 2. Following the Virtual Tutorial script, organize your Corpus and show the last **10** words using the `tail()` function. Be sure to change the arguments in `tail()` to display 10-words! [1pts]

```{r}


```

## 3. Filter out stop words from the `book` corpus include names such as `"Harry"`, `"Ron"`. and `"Hermione"` [1pts]

```{r}


```

## 4. Select the book you'd like to visualize as a word cloud and `subset()` the corpus to include that book. Create the wordcloud using the function: `wordcloud2()` [1pts]

**Note**: You may not select the 6th book for this visualization!

```{r, fig.align = "center", fig.width = 10, fig.height = 6}


```




## 5. The code below loads in a social network called "Zachary's karate club" and saves it into an variable called `karate`. Use the `plot()` function to visualize the karate club members. You do not need to apply any layout or additional functions for this [1pts]


```{r}
data("karate")
```

Plot goes below

```{r}

```

## 6. Calculate global metrics of the network and explain what each of them means [3pts]

- Density
- Average Path Length
- Diameter

```{r}
# Density

# Average Path Length

# Diameter

```

## 7. Calculate local metrics of the network (e.g., centrality measures) and explain [1pts]

- Degree
- Closeness
- Betweenness

```{r}
# Degree

# Closeness

# Betweenness

```

## 8. Who has the highest degree? Closeness? Betweenness?



# Part 2. Course Knowledge

## 1. Compare and contrast the 3 centrality measures for networks. How are they similar? How are they different? What do each tell us about a node in a network? Give an example in a work-environment of a person who has the highest `degree` and someone who has the highest `closeness`. Justify your answer [3pts]



## 2. In a high school setting, the president of the associated students organization may have the highest `degree` due to their connections with other students. Does this automatically mean they are also the `closest` or highest in `betweenness`? Why or why not? [2pts]




