---
title: "Assignment 03"
author: "Marwin Carmo"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

```{r}
library(ez) 
library(nlme)
```


```{r}
mydata <- read.csv("../data/mydata.csv")

# To run the ANOVA, I need to filter out some participants that don't have all three time points
mydata_clean <- mydata[!is.na(mydata$insomnia_severity), ]
timepoint_counts <- table(mydata_clean$record_id)
complete_ids <- names(timepoint_counts[timepoint_counts == 3])
mydata_complete <- mydata_clean[mydata_clean$record_id %in% complete_ids, ]
mydata_complete$redcap_event_name <- factor(mydata_complete$redcap_event_name)
```

## (1)	Select a variable in your data for modeling over time. (1 variable, at least 3 occasions). Use the same variable and data as Assignment 2.

Consistent with last week's assignment, I will work with the outcome of insomnia severity.

## (2)	Repeated Measures ANOVA  

### a.	Run repeated measures ANOVA using the `ezANOVA` function in the `ez` package

```{r}
m1 <- ezANOVA(data=mydata_complete, dv=insomnia_severity, wid=record_id, within=redcap_event_name)

```

### b.	Decide whether you use the univariate or multivariate test, and justify your decision

Because the Mauchly's test for sphericity is significant (W = `r round(m1[[2]]$W,2)`, *p* < 0.001), we assume there is a violation of heterogeneity of variances and use the multivariate test.

### c.	If you decide to use the univariate test, test for polynomial contrasts (e.g., linear, quadratic time effects) with proper time spacing using the aov function


### d.	If you decide to use the multivariate test, run the multivariate test (i.e., the MANOVA approach)

```{r}
# contrasts for unequally spaced time
contrasts(mydata_complete$redcap_event_name) <- contr.poly(c(0, 1.5, 6))

mUN <- gls(insomnia_severity ~ redcap_event_name, corr = corSymm(form = ~1|record_id), weights = varIdent(form = ~ 1 | redcap_event_name), 
          method="ML",data=mydata_complete)

mCS <- gls(insomnia_severity ~ redcap_event_name, corr = corCompSymm(form = ~1|record_id), method="ML", data=mydata_complete)

anova(mUN, mCS)
```


### e.	Write a few sentences reporting the results and their interpretation.

I examined how insomnia severity changed over three measurement occasions (baseline, posttreatment, and followup) in a sample of 191 participants. The assumption of equal variance and covariance, tested with the Mauchly's test of Sphericity did not adequately describe the data, as indicated by the significant p-value of the test. Therefore, I fitted a multivariate ANOVA model, in which each variance and covariance was estimated separately. The Unstructured model fitted significantly better than a model assuming Compound Symmetry ($\chi^2$(4) = 84.32, *p* < 0.001), and had lower AIC (3316.17 vs. 3392.48) and BIC (3355.32 vs. 3414.24). Significant mean differences in insomnia severity were observed across occasions within the multivariate model, F(1, 2) = 186.38, *p* < .001. The linear time  slope was significant (*p* < 0.001), with an estimated coefficient of -5.01 (SE = 0.28). The quadratic time slope was also significant and positive (*b* = 2.78, SE = 0.24, *p* < 0.001), indicating that while insomnia symptoms decreased over time, the rate of this improvement slowed down at later time points.

## (3)	Include the code you used to complete the assignment.

