---
title: "Week 1 - Introduction"
author: "Marwin Carmo"
format: 
  revealjs:
    scrollable: true
    #chalkboard: true
    incremental: true
    code-tools: true
    code-copy: true
    code-line-numbers: true
    code-link: true
    preview-links: true
    slide-number: true
    self-contained: true
    fig-height: 4
    fig-width: 6
    fig-align: center
    #css: custom.css
    #theme: psc290-23
    # highlight-style: atom-one-dark
    margin-left: "0"
    margin-right: "0"
    width: 1400
    # height: 900
    footer: "PSC 012Y - Data Visualization in the Social Sciences"
    logo: "https://github.com/emoriebeck/psc290-data-viz-2022/raw/main/01-week1-intro/02-code/02-images/ucdavis_logo_blue.png"
editor: source
editor_options: 
  chunk_output_type: console
---

## Who am I

-   2nd Year Graduate Student in Quantitative Psychology;

-   MSc in Psychiatry at University of Sao Paulo, Brazil;

-   Advised by Dr. Philippe Rast;

-   Studying intraindividual variability.

## Lab and homework dynamics

- All computations will use the R language;

- Assignments are released before Lab and should be completed during lab time, using the Lab computers;

- You will not be able to submit the assignment from your personal computer as we implemented IP filters;

- You will submit an `.rmd` file alongside a "knitted" document.

- Office hours on Monday 4:20-6:20PM via Zoom (link on Canvas)

- Questions via email at [mmcarmo@ucdavis.edu](mailto:mmcarmo@ucdavis.edu)

# Installing R and RStudio

## Download and Install R

- <https://cran.r-project.org/>

- **Windows**
  - Choose base and then click on Download R 4.4.1 for Windows.

- **macOS**
  - Select the latest release for your version of macOS and download the installer.
  - Open the `.pkg` file and follow the installation instructions.

- **Linux**
  - Follow the instructions for your distribution.
  
## Download and Install RStudio

- <https://posit.co/download/rstudio-desktop/>
- Choose the appropriate version for your operating system (Windows, macOS, or Linux).

## Setting Up R Packages

- After installing both R and RStudio, open RStudio on your computer.

-  Install the necessary packages by typing the following commands in the RStudio Console:

::: fragment
```{r, eval = FALSE, echo=TRUE}
install.packages(c("rmarkdown", "tidyverse", "knitr"))
```
:::

- Load each package with the `library()` function.

::: fragment
```{r, eval = FALSE, echo=TRUE}
library(rmarkdown)
library(tidyverse)
library(knitr)
```
:::

## Let's test it out

- Create a new temporary `.rmd` file from the drop down menu and knit it to a Word document
