---
title: "Assignment 08"
author: "Marwin Carmo"
output: 
  pdf_document:
    extra_dependencies: ["caption", "booktabs", "geometry", "longtable"]

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```


I specified a multivariate latent growth model to examine the trajectories of insomnia severity and anxiety from baseline to post-test. More specifically, whether initial levels of anxiety were related to change in insomnia severity. 

The model defined time-specific latent factors for insomnia and anxiety, each measured by seven items. The growth component for each construct was defined by an intercept factor (representing baseline levels, with loadings fixed at 1 for both time points on the respective latent factor) and a slope factor (representing the change from baseline to post-test, with loadings fixed at 0 for baseline and 1 for post-test). Given the ordered categorical nature of the items, I used the Diagonally Weighted Least Squares (DWLS) estimator.

The model demonstrated good overall fit to the data: $\chi^2$(338) = 510.29, *p* < 0.001; CFI = 0.965; TLI = 0.961; RMSEA = 0.051 (90% CI [0.042, 0.060]); SRMR = 0.080. Addressing the primary research question, the covariance between baseline anxiety levels  and the subsequent change in insomnia severity was not statistically significant ($\beta$ = -0.006, SE = 0.015, *p* = 0.687). This suggests that initial anxiety levels did not predict the magnitude of change in insomnia severity from baseline to post-test in this sample.



```{r message=FALSE}
library(lavaan)
library(tidyr)
library(dplyr)
```


```{r message=FALSE}
dat <- read.csv("../data/clean_data2.csv")

mydata <- dplyr::filter(dat, redcap_event_name %in% c(
    "elegibilidade_arm_1", 
    "desfechos_arm_1", "followup_arm_1")
  ) |> 
  dplyr::mutate(
    redcap_event_name = factor(dplyr::case_when(
      redcap_event_name == "elegibilidade_arm_1" ~ 1,
      redcap_event_name == "desfechos_arm_1" ~ 2,
      redcap_event_name == "followup_arm_1" ~ 3
    ))) |> 
  dplyr::select(record_id, redcap_event_name,
                dplyr::starts_with("igi"),
                dplyr::starts_with("ehad"))


wide_data <- mydata |> 
  tidyr::pivot_wider(
    id_cols = "record_id",
    names_from = "redcap_event_name",
    names_sep = ".",
    values_from = c(igi_1a:ehad_14)
  )
```

$$
\begin{longtable}{l r r r r}
\caption{Parameter Estimates for the Two-Time-Point Latent Growth Model of Insomnia and Anxiety} \label{tab:lgm2wavelong} \\ % Caption goes first for longtable
\toprule
Parameter Path                                  & Estimate (B) & SE    & z-value & p-value \\
\midrule
\endfirsthead % Header for the first page

$\chi^2$ (df)                                   & 510.288 (338) &       &         & <.001   \\
CFI                                             & 0.965         &       &         &         \\
TLI                                             & 0.961         &       &         &         \\
RMSEA [90\% CI]                                 & 0.051         & \multicolumn{3}{l}{[0.042, 0.060]} \\
SRMR                                            & 0.080         &       &         &         \\
\addlinespace
\multicolumn{5}{l}{\textbf{Measurement Model Loadings (Baseline, Time 1)}} \\
\textit{Insomnia Severity T1} $\rightarrow$       &               &       &         &         \\
\quad igi\_1a.1                                 & 1.000         & --    & --      & --      \\
\quad igi\_1b.1                                 & 1.006         & 0.112 & 8.973   & <.001   \\
\quad igi\_2.1                                  & 0.666         & 0.075 & 8.846   & <.001   \\
\quad igi\_3.1                                  & 1.188         & 0.126 & 9.398   & <.001   \\
\quad igi\_4.1                                  & 1.280         & 0.141 & 9.089   & <.001   \\
\quad igi\_5.1                                  & 1.082         & 0.118 & 9.162   & <.001   \\
\quad igi\_1c.1                                 & 0.589         & 0.093 & 6.316   & <.001   \\
\addlinespace
\textit{Anxiety T1} $\rightarrow$        &               &       &         &         \\
\quad ehad\_1.1                                 & 1.000         & --    & --      & --      \\
\quad ehad\_3.1                                 & 1.466         & 0.108 & 13.526  & <.001   \\
\quad ehad\_5.1                                 & 0.914         & 0.073 & 12.453  & <.001   \\
\quad ehad\_7.1                                 & 0.615         & 0.059 & 10.495  & <.001   \\
\quad ehad\_9.1                                 & 1.054         & 0.087 & 12.089  & <.001   \\
\quad ehad\_11.1                                & 1.129         & 0.094 & 11.956  & <.001   \\
\quad ehad\_13.1                                & 0.940         & 0.077 & 12.275  & <.001   \\
\addlinespace
\multicolumn{5}{l}{\textbf{Growth Factor Means}} \\
Insomnia Intercept (ItcI, Baseline)             & 0.000         & 0.021 & 0.000   & 1.000   \\
Anxiety Intercept (ItcA, Baseline)              & 0.000         & 0.018 & 0.000   & 1.000   \\
Insomnia Slope (SlopeI, Change T1-T2)           & 0.000         & 0.028 & 0.000   & 1.000   \\
Anxiety Slope (SlopeA, Change T1-T2)            & 0.000         & 0.024 & 0.000   & 1.000   \\
\addlinespace
\multicolumn{5}{l}{\textbf{Growth Factor (Co)Variances}} \\
\textit{Variances:}                             &               &       &         &         \\
\quad Var(ItcI)                                 & 0.140         & 0.032 & 4.431   & <.001   \\
\quad Var(ItcA)                                 & 0.166         & 0.019 & 8.588   & <.001   \\
\quad Var(SlopeI)                               & 0.086         & 0.022 & 3.962   & <.001   \\
\quad Var(SlopeA)            & 0.018         & 0.013 & 1.346   & 0.178   \\
\addlinespace
\textit{Covariances (Selected):}                &               &       &         &         \\
\quad Cov(ItcI, ItcA)                           & 0.106         & 0.013 & 8.205   & <.001   \\
\quad Cov(ItcA, SlopeI)                         & -0.006        & 0.015 & -0.403  & 0.687   \\
\quad Cov(ItcA, SlopeA)\textsuperscript{c}      & 0.041         & 0.014 & 2.972   & 0.003   \\
\quad Cov(SlopeI, SlopeA)\textsuperscript{c}    & 0.111         & 0.017 & 6.413   & <.001   \\
\addlinespace
\multicolumn{5}{l}{\textbf{Residual Variances of Latent Factors (Time-Specific)\textsuperscript{a}}} \\
Insomnia (In)                                   & 0.113         & 0.016 & 7.055   & <.001   \\
Anxiety (An)                                    & 0.055         & 0.009 & 6.314   & <.001   \\
\bottomrule
\end{longtable}
$$

```{r message=FALSE}
## Multivariate Linear LGM

MLlgm <- '
insomnia_severity1 =~ igi_1a.1 + igi_1b.1 + igi_1c.1 + igi_2.1 + igi_3.1 + igi_4.1 + igi_5.1
insomnia_severity2 =~ igi_1a.2 + igi_1b.2 + igi_1c.2 + igi_2.2 + igi_3.2 + igi_4.2 + igi_5.2

anxiety1 =~ ehad_1.1 + ehad_3.1 + ehad_5.1 + ehad_7.1 + ehad_9.1 + ehad_11.1 + ehad_13.1
anxiety2 =~ ehad_1.2 + ehad_3.2 + ehad_5.2 + ehad_7.2 + ehad_9.2 + ehad_11.2 + ehad_13.2

#Level/Intercept (all constraint to 1)
ItcI =~ 1*insomnia_severity1 + 1*insomnia_severity2  
ItcA =~ 1*anxiety1 + 1*anxiety2 

# Slope
SlopeI =~ 0*insomnia_severity1 + 1*insomnia_severity2 
SlopeA =~ 0*anxiety1 + 1*anxiety2 

#Residuals (Equality constraints)
insomnia_severity1 ~~ In*insomnia_severity1
insomnia_severity2 ~~ In*insomnia_severity2


anxiety1 ~~ An*anxiety1
anxiety2 ~~ An*anxiety2


#Intercept & slope means
ItcI~1
ItcA~1
SlopeI~1
SlopeA~1

#Intercept & slope variances
ItcI ~~ ItcI
ItcI ~~ ItcA
ItcI ~~ SlopeI
ItcI ~~ SlopeA
ItcA ~~ ItcA
ItcA ~~ SlopeI
ItcA ~~ SlopeA
SlopeI ~~ SlopeI
SlopeI ~~ SlopeA
SlopeA ~~ SlopeA
'

#creating a new object 'MLlgm' & running a SEM model
fit_MLlgm <- lavaan(MLlgm, data=wide_data, estimator = "DWLS",
                    auto.var=TRUE, auto.fix.first=TRUE, auto.cov.lv.x=TRUE,
                    meanstructure = TRUE, int.ov.free = TRUE)

#getting summary of model estimates (fit & parameters)
summary(fit_MLlgm, fit.measures=TRUE)
```

```{r eval=FALSE, include=FALSE}
## Multivariate Linear LGM

MLlgm <- '
insomnia_severity1 =~ igi_1a.1 + igi_1b.1 + igi_1c.1 + igi_2.1 + igi_3.1 + igi_4.1 + igi_5.1
insomnia_severity2 =~ igi_1a.2 + igi_1b.2 + igi_1c.2 + igi_2.2 + igi_3.2 + igi_4.2 + igi_5.2
insomnia_severity3 =~ igi_1a.3 + igi_1b.3 + igi_1c.3 + igi_2.3 + igi_3.3 + igi_4.3 + igi_5.3

anxiety1 =~ ehad_1.1 + ehad_3.1 + ehad_5.1 + ehad_7.1 + ehad_9.1 + ehad_11.1 + ehad_13.1
anxiety2 =~ ehad_1.2 + ehad_3.2 + ehad_5.2 + ehad_7.2 + ehad_9.2 + ehad_11.2 + ehad_13.2
anxiety3 =~ ehad_1.3 + ehad_3.3 + ehad_5.3 + ehad_7.3 + ehad_9.3 + ehad_11.3 + ehad_13.3

#Level/Intercept (all constraint to 1)
ItcI =~ 1*insomnia_severity1 + 1*insomnia_severity2 + 1*insomnia_severity3 
ItcA =~ 1*anxiety1 + 1*anxiety2 + 1*anxiety3

# Slope
SlopeI =~ 0*insomnia_severity1 + 1*insomnia_severity2 + 4*insomnia_severity3
SlopeA =~ 0*anxiety1 + 1*anxiety2 + 4*anxiety3

#Residuals (Equality constraints)
insomnia_severity1 ~~ In*insomnia_severity1
insomnia_severity2 ~~ In*insomnia_severity2
insomnia_severity3 ~~ In*insomnia_severity3

anxiety1 ~~ An*anxiety1
anxiety2 ~~ An*anxiety2
anxiety3 ~~ An*anxiety3

#Intercept & slope means
ItcI~1
ItcA~1
SlopeI~1
SlopeA~1

#Intercept & slope variances
ItcI ~~ ItcI
ItcI ~~ ItcA
ItcI ~~ SlopeI
ItcI ~~ SlopeA
ItcA ~~ ItcA
ItcA ~~ SlopeI
ItcA ~~ SlopeA
SlopeI ~~ SlopeI
SlopeI ~~ SlopeA
SlopeA ~~ SlopeA
'

fit_MLlgm <- lavaan(MLlgm, data=wide_data, estimator = "DWLS", std.lv = TRUE,
                    auto.var=TRUE, auto.fix.first=TRUE, auto.cov.lv.x=TRUE,
                    meanstructure = TRUE, int.ov.free = TRUE)


summary(fit_MLlgm, fit.measures=TRUE)
```

```{r eval=FALSE, include=FALSE}

## Multivariate Linear LGM

MLlgmQ <- '

insomnia_severity1 =~ igi_1a.1 + igi_1b.1 + igi_1c.1 + igi_2.1 + igi_3.1 + igi_4.1 + igi_5.1
insomnia_severity2 =~ igi_1a.2 + igi_1b.2 + igi_1c.2 + igi_2.2 + igi_3.2 + igi_4.2 + igi_5.2
insomnia_severity3 =~ igi_1a.3 + igi_1b.3 + igi_1c.3 + igi_2.3 + igi_3.3 + igi_4.3 + igi_5.3

anxiety1 =~ ehad_1.1 + ehad_3.1 + ehad_5.1 + ehad_7.1 + ehad_9.1 + ehad_11.1 + ehad_13.1
anxiety2 =~ ehad_1.2 + ehad_3.2 + ehad_5.2 + ehad_7.2 + ehad_9.2 + ehad_11.2 + ehad_13.2
anxiety3 =~ ehad_1.3 + ehad_3.3 + ehad_5.3 + ehad_7.3 + ehad_9.3 + ehad_11.3 + ehad_13.3

#Level/Intercept (all constraint to 1)
ItcI =~ 1*insomnia_severity1 + 1*insomnia_severity2 + 1*insomnia_severity3 
ItcA =~ 1*anxiety1 + 1*anxiety2 + 1*anxiety3

# Linear slope
SlopeI =~ 0*insomnia_severity1 + 1.5*insomnia_severity2 + 6*insomnia_severity3
SlopeA =~ 0*anxiety1 + 1.5*anxiety2 + 6*anxiety3

# Quadratic slope
QSlopeI =~ 0*insomnia_severity1 + 2.25*insomnia_severity2 + 36*insomnia_severity3
QSlopeA =~ 0*anxiety1 + 2.25*anxiety2 + 36*anxiety3

#Residuals (Equality constraints)
insomnia_severity1 ~~ In*insomnia_severity1
insomnia_severity2 ~~ In*insomnia_severity2
insomnia_severity3 ~~ In*insomnia_severity3

anxiety1 ~~ An*anxiety1
anxiety2 ~~ An*anxiety2
anxiety3 ~~ An*anxiety3

#Intercept & slope means
ItcI~1
ItcA~1
QSlopeI~1
QSlopeA~1

#Intercept & slope variances
ItcI ~~ ItcI
ItcI ~~ ItcA
ItcI ~~ SlopeI
ItcI ~~ SlopeA
ItcI ~~ QSlopeI
ItcI ~~ QSlopeA
ItcA ~~ ItcA
ItcA ~~ SlopeI
ItcA ~~ SlopeA
ItcA ~~ QSlopeI
ItcA ~~ QSlopeA
SlopeI ~~ SlopeI
SlopeI ~~ SlopeA
SlopeI ~~ QSlopeI
SlopeI ~~ QSlopeA
SlopeA ~~ SlopeA
SlopeA ~~ QSlopeI
SlopeA ~~ QSlopeA
QSlopeI ~~ QSlopeI
QSlopeI ~~ QSlopeA
QSlopeA ~~ QSlopeA
'


#creating a new object 'MLlgm' & running a SEM model
fit_MLlgmQ <- lavaan(MLlgmQ, data=wide_data, estimator = "DWLS",
                     auto.var=TRUE, auto.fix.first=TRUE, auto.cov.lv.x=TRUE,
                    meanstructure = TRUE, int.ov.free = TRUE)

#getting summary of model estimates (fit & parameters)
summary(fit_MLlgmQ, fit.measures=TRUE)
```


