---
title: "Lab 6: Multiple Comparisons & Factorial ANOVA"
subtitle: "PSC 103B"
author: "Marwin Carmo"
format: 
  revealjs:
    scrollable: true
    incremental: true
    code-tools: true
    code-copy: true
    code-line-numbers: true
    code-link: true
    preview-links: true
    slide-number: true
    self-contained: true
    fig-height: 4
    fig-width: 6
    fig-align: center
    margin-left: "0"
    margin-right: "0"
    width: 1400
    callout-icon: false
    # height: 900
    footer: "PSC 103B - Statistical Analysis of Psychological Data"
    logo: "https://github.com/emoriebeck/psc290-data-viz-2022/raw/main/01-week1-intro/02-code/02-images/ucdavis_logo_blue.png"
editor: source
execute:
  echo: true
editor_options: 
  chunk_output_type: console
---

## Read in data

For today's class, we're going to use the same NPAS dataset from Lab 5

```{r}
library(tidyverse)
library(lm.beta)
```

```{r}
nerdy <- read_csv("NPAS_Lab5.csv")
```

## Covariance and Correlation

- Up until now, when we have worked with more than one variable it has typically been that one of the variables is categorical, and the other is continuous.

- Covariance and correlation are summary statistics that quantify the relation between two (or more!) continuous (or ordinal) variables without making any statement about the direction.

## Covariance

- Recall 
