---
title: "Missing Data"
author: "PR"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
---

```{r, message = FALSE, echo = FALSE}
## Load packages
library( ggplot2 )
library( rstan )
# use multiple cores
options(mc.cores =  parallel::detectCores())
# compile model and save it 
rstan_options(auto_write = TRUE)
```
# Data

Data are from a recent study (March -- April 2020) on 86 patients infected with COVID-19 who were treated in the emergency department (ED) of Lausanne University Hospital, Switzerland. The first google hit on https://datasetsearch.research.google.com/
landed me on their page (I only noted afterwards that data are from Switzerland - nothing personal): https://zenodo.org/record/3763421

These data include a number of variables as well as a an outcome 7 days into entering the study. 
Here, we will use `fu7_death` as the outcome variable -- i.e., this variable records whether that study participant was alive or deceased by day 7.

The dataset is not really documented, but this is my guess of what these variables mean:
Outcome at day 7

 - `fu7_icu`: 1 = Intensive Care (lots of missings)
 - `fu7_iot`: 1 = Intubated (lots of missings)
 - `fu7_death`: 1 = Dead 

You are free to play around with these variables. 

The Other variables seem to have been recorded at the _entry_ to the study 7 days earlier and most of them are self explanatory. We will use some of them as predictors.

### Read in data 
Let's read in the data and peak into it 
```{r}
lus <- read.csv( file = "https://zenodo.org/record/3763421/files/Clinical_data.csv?download=1" )
head( lus, n =  5 )
```
### Subset Data 
Slect some variables of interest -- note that some have lot's of missing values. That is, I selected mostly those with few missings.
So far, we haven't talked about ways to deal with missing values, hence, I dropped all missing values with `na.omit()`.

I also recoded sex to by dummy-coded with 0 = male, and 1 = female

```{r}
lus2 <- lus[  , c("sex", "age",  "hbp", "diabetes", "heart_ischemia",
                                                 "asthma", "cp_bp_systolic",  "cp_bp_diastolic",
                                                 "cp_saturation", "creatine", "leukocytes",
                                                 "status_pulm", "fu7_death")] 
## Recode sex, to 0 = males, 1 = females
lus2$sex <- lus2$sex - 1

lus2


## We're left with N = 61
nrow(lus2)
```

# Question 1 

We predict `fu7_death` as this variable seems to have relatively few missings.

###  Formal Model

We are trying to predict the outcome `fu7_death` which is 0/1 coded. 0 is alive, 1 is dead.
A logistic regression seems to be a natural choice where some of the predictors will increase or decrease the odds of dying due to COVID related complications. 

As discussed in PSC204B, logistic regressions are general linear models based on a Bernoulli distribution with a logit link $\log\left( \frac{p}{1 - p} \right)$. In terms of defining this model, we can write 
\begin{align*}
\text{Likelihood:} \hspace{1cm}	y &\sim Bern( \theta ) = 
		\begin{cases}
			\theta \text{ if } y = 1\\
			1 - \theta  \text{ if } y = 0
		\end{cases}
	\text{ where } \theta \in \{0,1\} \\
& logit( \theta ) = log\left(\frac{\theta}{1-\theta}\right) = \beta_0 + x_1\beta_1 + \ldots + x_p\beta_p =  \mathbf{X}\boldsymbol{\beta} \\
\text{ Prior for all elements in }\boldsymbol{\beta}\text{:} \hspace{1cm} \boldsymbol{\beta} &\sim \mathcal{N}(\mathbf{0}, 3\mathbf{I})
\end{align*}

In words, our likelihood is defined as a Bernoulli function. Given the logit link that removes all bounds from our linear model, we can use relatively uninformed normal distributions for each of the $\beta$'s. Here, I chose a normal distribution with mean of zero and a standard deviation of 3 for each of the $\beta$'s

The next step is to write this model up in `stan`. 

# Question 2

Now, take `sex`, `age`, and `heart_ischema` as predictors and `fu7_death` as the outcome.

a) Create a `dat` list object that takes all the arguments you defined in the `data {}` block of your stan model
b) Sample from the model but first _without_ the likelihood and find a way to generate a meaningful prior predictive plot
c) After that: Sample from the posterior and 
	1) print the output and report interesting findings in terms of odds, 
	2) plot the posterior predictive distribution. 
 
```{r}
library( rstan )
```


### Hint 3: Results and _Posterior_ PD
```{r}
## library( brms )
## mod0 <- brm(fu7_death | mi() ~ sex + age + heart_ischemia,
##             data = lus2)
## stancode(mod0 )


X <- cbind(1, lus2$sex, lus2$age, lus2$heart_ischemia )

## Predictor contains one missing: We replace this with zero (we could esimate this with mice)
## row 40 contains missing - set to 0
X[40, 4] <- 0
X

mis_index <- ifelse( is.na(lus2$fu7_death), 1, 0)
mis_index
N_mis <- sum(mis_index )
N_mis

N_obs <- length(mis_index ) - N_mis
N_obs

pos_missing <- which( mis_index == 1 )
pos_missing

## replace lus2$fu7_death NA's with 0
lus2$fu7_death[is.na( lus2$fu7_death )] <- 0
lus2$fu7_death

dat <- list(N = nrow( X ),
            N_mis = N_mis,
            pos_missing =  pos_missing,
            p = ncol( X ),
            X =  X,
            y =  lus2$fu7_death)

model1 <- stan_model(file = "./missing.stan" )
#model1 <- stan_model(file = "./missingness.stan" )

fit1 <- sampling( model1, data = dat, iter =  3000 )

## Show estimated 
print( fit1, pars =  c("theta_mis"), probs=c(.05, .95) )
print( fit1, pars =  c("beta"), probs=c(.05, .95) )

plot(fit1, pars =  c("theta_mis") )


posterior::as_draws(fit1)
```



```{r}
kidiq <- haven::read_dta("http://www.stat.columbia.edu/~gelman/arm/examples/child.iq/kidiq.dta")

kidiq

library(mice)
set.seed(1955)
kidiq100_mar <- ampute(kidiq, prop = 0.5, 
                       pattern = data.frame(kid_score = c(0, 0, 0), 
                                            mom_hs = c(1, 1, 0), 
                                            mom_work = c(1, 1, 1), 
                                            mom_age = c(1, 1, 1), 
                                            mom_iq_c = c(1, 0, 1)), 
                       freq = c(.2, .4, .4), 
                       mech = "MAR")

kidiq100_mar
kidiq100_mar <- kidiq100_mar$amp
```


```{r}
# Recode mom_hs to factor

kidiq100_mar$mom_hs

kidiq100_mar$mom_hs <- factor(kidiq100_mar$mom_hs, labels = c("no", "yes"))

md.pattern(kidiq100_mar, rotate.names = TRUE)


library(brms )
head(kidiq100_mar )

listwdel <- brm(kid_score ~ mom_iq + mom_hs, data = kidiq100_mar)
listwdel

mar <- brm(kid_score | mi() ~ mom_iq + mom_hs, data = kidiq100_mar)
stancode(mar )
mar

kidiq$mom_hs <- factor(kidiq$mom_hs, labels = c("no", "yes"))
full <- brm(kid_score  ~ mom_iq + mom_hs, data = kidiq)
full



```
