---
title: "PSC 204B Homework 1"
date: "Due Date: January 19, 2024"
output:
  pdf_document:
    toc: no
---

```{r echo = F, warning = F, message = F}
# The following packages are needed to complete this homework assignment. The following code checks whether the packages are already installed, if not they will be installed. 
if (!require("dplyr")) install.packages("dplyr")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("ggpubr")) install.packages("ggpubr")
if (!require("psych")) install.packages("psych")
if (!require("DescTools")) install.packages("DescTools")
```

```{r echo = F, message = F}
# Load the required packages 

library(dplyr)
library(ggplot2)
library(ggpubr)
library(psych)
library(DescTools)
```

```{r echo = F}
# The data you will be using for this Homework Assignment is from the study below:
 
# Kim, S. E., Kim, H. N., Cho, J., Kwon, M. J., Chang, Y., et al. (2016) Correction: Direct and indirect effects of five factor personality and gender on depressive symptoms mediated by perceived stress. *PLOS ONE, 11*: e0157204.

# The data file can be found on Canvas under the Assignments tab -> Week 2 -> Homework -> hw2data.csv

# The `hw2data.csv` file contains the following variables:

#  Stress: Total perceived stress score from self-reported stress questionnaire 
#  CESD: Total depression score for the Center for Epidemiological Studies Depression Scale
#  N: Total score on Neuroticism from the Revised NEO Personality Inventory
#  E: Total score on Extroversion from the Revised NEO Personality Inventory
#  O: Total score on Openness to Experience from the Revised NEO Personality Inventory
#  A: Total score on Agreeableness from the Revised NEO Personality Inventory
#  C: Total score on Contentiousness from the Revised NEO Personality Inventory
#  sex: Binary variable representing biological sex (0 = male; 1 = female)
```

```{r echo = F, message = F}
# Read in the data 

hw1 = read.csv("hw1data.csv", header = TRUE)
```

# Question 1 **(2 points)**

Fit a simple linear regression model using biological sex (*sex*; 0 = male; 1 = female) to predict depression (*CESD*). Write out the predicted regression equation using R markdown equation notation in the space below (see the lab), and interpret each parameter (see the lab for what is expected). Round all numbers to two decimal places.

```{r message = F}
## Code
mod1 <- lm(CESD ~ sex, data = hw1)
summary(mod1)
```

-   **Equation**: $\widehat{Depression_i} = 9.79 + 2.45 \times Sex_i$
-   **(Intercept)**: The expected value of depression is 9.79 when sex is male.
-   **Sex**: Females are predicted to have 2.45 more units of depression than males. This slope is significantly different from 0, indicating that Depression and sex are significantly associated.

# Question 2 **(6.5 points)**

## Part a) **(2.5 points)**

Fit a simple linear regression model using neuroticism (*N*) to predict stress (*Stress*). Write out the predicted regression equation and interpret each of the parameters found in the regression model.

```{r message = F}
## Code
mod2 <- lm(Stress ~ N, data = hw1)
summary(mod2)
```

-   **Equation**: $\widehat{Stress_i} = -2.66 + 0.36 \times Neuroticism_i$

-   **(Intercept)**: The expected value of stress is -2.27 when Neuroticism is equal 0.

-   **N**: An one-unit increase in Neuroticism is associated with a 0.36 increase in Stress. This slope is significantly different from 0, indicating that Stress and Neuroticism are significantly associated.

## Part b) **(1 point)**

Plot the model diagnostic plots, and assess whether the assumptions of linearity, homogeneity of variances, and normality of residuals are met. Explain your reasoning. 

```{r, message = FALSE}
plot(mod2)
```

- **Linearity**: The predicted line of the Residuals vs. Fitted plot is expected to be flat to indicate zero correlation between the predicted and residual values. The assumption of linearity in this model is well met.

- **Homogeneity of Variances**: We can also use the Residuals vs. Fitted plot to investigate the homogeneity of variances. Different from the random pattern expected when the variance is homogeneous, we observe a peculiar pattern, indicating heterocedasticity.

- **Normality of Residuals**: The normality of residuals can be assessed with the QQ Plot. All of the points are expected to fall on the diagonal line. In this model, we observe an expressive departure from normality for the residuals since the standardized residuals greater than zero show a positive slope, deviating from the reference dotted line. 

## Part c) **(2.5 points)**

Repeat the regression analysis above in Part a, but using mean-centered Neuroticism (*N*) to predict *Stress*. Call the mean-centered Neuroticism variable *N_c*. Write out the predicted regression equation and interpret each of the parameters found in the regression model. **In each of your interpretations, comment on whether the parameter changed from the original analysis, and why or why not.**

```{r message = F}
## Code
hw1$N_c <- scale(hw1$N, scale = FALSE)

mod3 <- lm(Stress ~ N_c, data = hw1)
summary(mod3)
```

-   **Equation**: $\widehat{Stress_i} = 17.44 + 0.36 \times Neuroticism_{c_i}$

-   **(Intercept)**: The predicted value of stress is 17.44 when Neuroticism at its mean level. By centering the predictor variable the value of the intercept also changed because now we've changed the reference point for the Neuroticism.

-   **N**: The predicted value of $N_c$ does not differ from the predicted value of $N$ because centering the preditor does not alter its linear relationship with the outcome. It still means that an one-unit increase in Neuroticism is associated with a 0.36 increase in Stress. The statistical significance of the slope also did not change.

## Part d) **(0.5 points)**

Using the model from part (b), what is the predicted score for someone who scores 1.5 points *above the mean*?

```{r}
# Code

17.44 + 0.36 * 1.5

```




# Question 3 **(1.5 points)**

Create two scatter plots to illustrate the relation between Stress and Neuroticism in the two analyses performed in Question 2. Arrange the graphs in two rows. Stress should be on the y axis in both graphs. In the first row, show the relationship between Stress and uncentered Neuroticism (*N*). In the second row, show the relationship between Stress and mean-centered Neuroticism (*N_c*). On each graph, add a line of best. Be sure to label your axes, and make sure that all graphs have the same x-axis and y-axis range.


```{r message = F}
## Code

# Original, uncentered Data
g1 <- ggplot(data = hw1, aes(y = Stress, x = N)) + 
  geom_point() +
  theme_classic() +
  xlab('Neuroticism') +
  ylab('Stress') +
  geom_smooth(method = 'lm', se = F) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(-25, 35))

# Mean-centered FNE
g2 <- ggplot(data = hw1, aes(y = Stress, x = N_c)) +
  geom_point() +
  theme_classic() +
  xlab('Centered Neuroticism') +
  ylab('Stress') +
  geom_smooth(method = 'lm', se = F) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  coord_cartesian(xlim = c(-25, 35))
  
ggarrange(g1, g2, nrow = 2)
```

# Extra Credit **(1 point)**

Conduct a regression to assess whether min-centered Neuroticism (*N_min*) predicts **median-centered** *Stress* (*Stress_med*). In the space below, report and interpret the intercept. Keep in mind that the value of the intercept is now relative to something since it has been centered. 

```{r message = F}
## Code

hw1$N_min <- hw1$N - min(hw1$N)
hw1$Stress_med <- hw1$Stress - median(hw1$Stress)

mod4 <- lm(Stress ~ N_min, data = hw1)
summary(mod4)

```

-   **Report the value of the intercept**: -10.11

-   **Interpret the intercept**: The intercept now represents the predicted median of the outcome variable when the predictor variable is at its minimum value.

```{r}
ggplot(data = hw1, aes(y = Stress_med, x = N_min)) +
  geom_point() +
  theme_classic() +
  xlab('Centered Neuroticism') +
  ylab('Stress') +
  geom_smooth(method = 'lm', se = F) +
  geom_vline(xintercept = 0, linetype = "dashed") #+ coord_cartesian(xlim = c(-25, 35))
```

