---
title: "Homework 3"
date: "Winter 2024"
output:
  #word_document: default
  pdf_document: default
---

```{r echo = F, message = F, warning = F}
library(ggplot2)
library(dplyr)
library(knitr)
library(broom)
library(flextable)
library(stargazer)
```

```{r echo = F, message = F,  warning = F}
# For this homework, you will use the Age Religion Health data set.
arh <- read.csv("arh.csv")

# To re-code the smoke01 variable (run all this code before doing the assignment)
arh <-
  arh %>%
  mutate(smoke01 = 
           ifelse(is.na(smoke01),
                  NA,
                  ifelse(smoke01 == 1,
                         "Smoker",
                         "Non-Smoker")))
```

```{r echo = F}
# IMPORTANT: Whenever you interpret a slope or intercept from a regression model output, please make sure to explicitly mention group names, when applicable. 

# Do not simply say "The intercept is the predicted score when XXX is 0". You must say what it means for XXX to be 0. Be as specific as possible.  

# Do not simply say "The slope is the expected change for every one unit change in XXX". You must say what a one-unit change in XXX means. Be as specific as possible.
```

# Question 1

## Part a)

Using dummy codes, run a regression to assess whether *depress2001* differs by marriage status (*married01*). Use "Married" as the reference group. Create a table to display the results of the analysis (Hint: try using the `bind_rows()` , `tidy()` , and `kable`, or `stargazer`' functions. You could also use any other functions that you would like, as long as the table is clean and clear). Then, in the space below, write the regression equation and interpret the overall model (was there a significant effect of *married01* on *depress2001*? Cite the appropriate statistics) and the intercept.

```{r message = F, warning = F, warning = F,  results = 'asis'}

arh_q1 <- arh |> 
  dplyr::mutate(divorced_v_married = ifelse(married01 == "Divorced", 1, 0),
                nevmarried_v_married = ifelse(married01 == "Never Married", 1, 0),
                separated_v_married = ifelse(married01 == "Separated", 1, 0),
                widowed_v_married = ifelse(married01 == "Widowed", 1, 0),
                )

mod1a <- lm(depress2001 ~ divorced_v_married + nevmarried_v_married + separated_v_married + widowed_v_married, data = arh_q1)
stargazer(mod1a, type = "latex", header = FALSE, ci=TRUE, digits=2, title = "Depression predicted by Marriage Status.")
```

$$ \widehat{\texttt{depress2001}_i} = 11.39 + 1\texttt{Divorced}_i + 1.68\texttt{NevMarr}_i + 0.012\texttt{Separated}_i + 1.38\texttt{Widowed}_i$$

**Interpretations**:

-   **Overall model**: The overall model is statistically significant (*F*(4, 817) = 4.0, *p* = 0.003), meaning that marriage status has an effect on Depression. However, the model only explains a small amount of the variance in the dependent variable ($R^2_{adj}$=0.14).

-   **Intercept**: The coefficient for the intercept is 11.39 and statistically significant (*p*<.001), which means that the predicted value of Depression for a person who is Married is 11.39.

## Part b)

Based on the regression performed above in Part A, which groups can you conclude are significantly different (at *p* \< .05)? List each specific pair of groups that differed.

**Answer**:

-   Widowed vs. Married: *b* = 1.38, *p* < 0.001.

## Part c)

Create a plot with *married01* on the x-axis and average levels of *depress2001* on the y-axis (i.e., using `geom_point()`). Include standard error bars for plus and minus 1 SE.

```{r message = F, warning = F}

arh_mean <- arh |> 
  dplyr::with_groups(
    married01, 
    summarise,
    mean = mean(depress2001, na.rm = TRUE),
    n = length(depress2001),
    sd = sd(depress2001, na.rm = TRUE),
    se = sd / sqrt(n),
    upper = mean + se,
    lower = mean - se
    ) |> 
  dplyr::filter(!is.na(married01))

arh_mean |> 
  ggplot(aes(x = married01, y = mean)) +
  geom_point(size = 5) +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  theme_minimal() +
  labs(x = "Marriage status", y = "Depression")

```

# Question 2

## Part a)

Repeat the regression from Question 1, Part A (assessing differences in *depress01* based on *married01* grouping), but this time use "Divorced" as the reference group. Create a table to display the results of the analysis (Hint: try using the `bind_rows()` , `tidy()` , and `kable` functions. You could also use any other functions that you would like, as long as the table is clean and clear).

```{r message = F, warning = F, results='asis'}
arh_q2 <- arh |> 
  dplyr::mutate(married_v_divorced = ifelse(married01 == "Married", 1, 0),
                nevmarried_v_divorced = ifelse(married01 == "Never Married", 1, 0),
                separated_v_divorced = ifelse(married01 == "Separated", 1, 0),
                widowed_v_divorced = ifelse(married01 == "Widowed", 1, 0),
                )

mod2a <- lm(depress2001 ~ married_v_divorced + nevmarried_v_divorced + separated_v_divorced + widowed_v_divorced, data = arh_q2)
stargazer(mod2a, type = "latex", header = FALSE, ci=TRUE, digits=2, title = "Depression predicted by Marriage Status.")
```

## Part b)

Interpret the slope for the dummy code for "Never Married" in both regressions (the one with "Married" as the reference group, and the one with "Divorced" as the reference group).

**Answer**:

-   **"Married" Reference Group**: In this mode, the slope for "Never Married" is statistically non-significant and positive (*b* = 1.68, 95% CI [-0.11, 3.46], *p* = 0.065). It represents the difference between the mean level of depression of the "Married" group and the mean depression of the "Never Married" group.

-   **"Divorced" Reference Group**: In this mode, the slope for "Never Married" is statistically non-significant and positive (*b* = 0.68, 95% CI [-1.41, 2.77], *p* = 0.524). It represents the difference between the mean level of depression of the "Divorced" group and the mean depression of the "Never Married" group.

## Part c)

Based on the analysis performed with "Divorced" as the reference group, which groups can you conclude are significantly different at *p* \< .05?

**Answer**:

-   None of the other groups are significantly different from the "Divorced" group.

# Question 3

## Part a)

Using dummy codes, run a two-way factorial ANOVA model with *self_worth2001* as the outcome and *married01*, *smoke01*, and the interaction between these variables as predictors. For the *married01* variable, use "Married" as the reference group; for the *smoke01* variable, use "Non-Smoker" as the reference group. Display the results of the analysis in a table in the same way as above (i.e., with `tidy()` and `kabl()`).

```{r message = F, warning = F, results='asis'}

arh$married01 <- factor(arh$married01, levels = c("Married", "Widowed", "Never Married", "Divorced", "Separated"))
arh$smoke01 <- factor(arh$smoke01, levels = c("Non-Smoker", "Smoker"))

mod3 <- lm(self_worth2001 ~ married01 * smoke01, data=arh)
stargazer(mod3, type = "latex", header = FALSE, ci=TRUE, digits=2, title = "Depression predicted by Marriage Status.")
```

## Part b)

Interpret the intercept.

-   **Answer**: WRITE ANSWER HERE

## Part c)

Among the non-smoker group, which group(s) are significantly different from the reference group, at *p* \< .05?

-   **Answer**: WRITE ANSWER HERE

## Part d)

Create a graph showing the means of *self_worth2001* (on the y axis) broken down by *married01* and *smoke01*, with +/- 1 SE bars. Try different ways of graphing the data (e.g., try putting *married01* on the x axis and coloring by *smoke01*, then try it the other way around; you could also try faceting the graph by *married01* and *smoke01*), and use the combination that you think emphasizes the most interesting effects in the data. Include a brief explanation of why you chose the graphed the data the way you did. There is not necessarily one right way to do this, as long as you can justify your decision.

```{r message = F, warning = F}


```

-   **Explanation**: WRITE ANSWER HERE
