---
title: "HW 7"
subtitle: Propensity Score Matching
output: 
  pdf_document:
    toc: true
    highlight: kate
toccolor: blue
urlcolor: blue
---

The data for this homework is stored in the `hbo.csv` file and contains information on tv shows, movies, etc. on HBO up to 2020. The variables we will be using today are:

* `type`: Whether the media is a TV show or a Movie/Other (factor)
* `rating`: The parental guidelines rating (e.g., G, PG, etc.; factor)
* `imdb_score`: The media's score on IMDB (numeric)
* `genre`: The media's genre (factor)

```{r message=FALSE, warning=FALSE}
# Load required packages
library(dplyr)
library(MatchIt)
library(stargazer)

# Read in data, convert character vars to factors,
# and filter to years of interest
hbo <- read.csv("hbo.csv") |>
  select(-X) |>
  mutate(across(where(is.character), as.factor)) |>
  filter(year >= 1999)
```


# Question 1 [1 pt.]

Recent decades have been talked about as a ["golden age" of television shows](https://en.wikipedia.org/wiki/Golden_Age_of_Television_(2000s%E2%80%93present)). Inspired by this, we will examine whether or not TV shows (from 1999 to 2020) are rated more highly than other types of media on HBO.

With the `hbo` data, use a linear regression to assess the effect of `type` on `imdb_score`. Based on the results of the regression, state below whether TV shows are rated significantly more highly than other types of media, at the $p < .05$ level.


```{r echo = T, warning = F, message = F}

```

* Answer here


# Question 2

## Part a [3 pts.]

We can't randomly assign media to be either a TV show or a movie, so let's redo the above analysis using a matched sample (use the functions from the `MatchIt` package). Use nearest neighbor matching, based on `rating` and `genre` (missing data for these variables were already filtered out for you, so you do not need to worry about that).

Display the matched regression output below.

```{r echo = T, warning = F, message = F}

```



## Part b [1 pt.]

When matched on `rating` and `genre`, do TV shows appear to be more highly rated than other types of media, at the $p < .05$ significant level? How does this compare to the original analysis above?

* Answer here


# Question 3

## Part a [3 pts.]

Re-do the propensity matched analysis, but this time use exact matching instead of nearest neighbor matching. Display the regression output below.

```{r}

```


## Part b [1 pt.]

Did any subgroups get dropped?

```{r echo = T, warning = F, message = F}

```

* Answer here

# Question 4 [1 pt.]

Using the `stargazer()` function, create a table of the regressions in questions 1, 2 and 4. (1 pt.)

```{r, results='asis'}

```

