---
title: "PSC 204B - Winter 2024 - Homework 6"
author: '"Your Name"'
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
subtitle: Multilevel Modeling
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message = FALSE}
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(foreign)
library(haven)
library(performance)
library(lme4)

```

We will continue to use the *popularity* dataset we used in lab, as well as a new dataset for a repeated measures example. The second dataset comes from Chapter 10 of *Longitudinal Analysis* by Lesa Hoffman, and contains data from 207 older adults without a dementia diagnosis who were assessed roughly every two years for eight years on their memory (*prose recall*). 

```{r}
popular = read_spss("popular2.sav")

memory = read_sas("SAS_Chapter10a.sas7bdat")
```

# Question 1

Use the *popularity* dataset to answer the following questions. In these questions, we are interested in predicting **teacher-rated** popularity of students (*popteach*) based on their gender. 

## Part a) [1 point]

Calculate the intra-class correlation coefficient for how much variability in teacher-rated popularity there is across classrooms. Report the value, and state whether this is considered a large amount of variation (provide a reason, based on cut-offs provided in lecture). 
```{r}
int_model <- lmer(popteach ~ 1 + (1|class), data= popular)
performance::icc(int_model)
```

**ICC Value:** 0.343
**Is this substantial?**: Since $\rho_{ic}$ > .05, the errors can be considered substantial.

## Part b) [1.5 point]

Run 3 regression models, each predicting teacher-rated popularity using gender, and report the output:

- A OLS regression model
- A multilevel model with a random intercept only
- A multilevel model with a random intercept and random slope

```{r}
ols_mod <- lm(popteach ~ sex, data = popular)
ml_int_mod <- lmer(popteach ~ sex + (1|class), data = popular)
ml_sp_mod <- lmer(popteach ~ sex + (1 + sex|class), data = popular)

summary(ols_mod)
summary(ml_int_mod)
summary(ml_sp_mod)
```

## Part c) [1 point]

How do the 3 models differ in terms of what is / is not allowed to differ across classrooms? In other words, what does it mean to include no random effects, a random intercept only, or a random intercept and random slope?

In the OLS model, the relationship between gender and popularity is assumed to be identical across all classrooms. That is, it gives us the average effect of gender on popularity. When we include a random intercept for classes, we allow them to have different average popularity levels, but thee effect of gender on popularity is still assumed to be the same. In the model with random intercepts and random slopes, classrooms can have different average popularity levels and the effect of gender on popularity is allowed to vary across classrooms.

## Part d) [1 point]

Report and interpret the fixed effects from the model with the random intercept and random slope. 

**(Intercept)**: The intercept represents the average level of teacher-rated popularity when the student is a boy.

**Effect of sex**: 

## Part e) [1 point]

How do the fixed effects and standard errors differ between the OLS model and the model with a random intercept and random slope?

**Fixed effect of intercept:**

**SE of intercept:**

**Fixed effect of gender:**

**SE of gender:**

# Question 2

## Part a) [1 point]

Create a plot to examine how prose recall (*recall*) changes with age (*tvage*), with age on the x-axis and prose recall on the y-axis. Be sure that observations from the same individual are connected with a line. Be sure to label your axes appropriately. 

```{r}
# code here
```


## Part b) [2 point]

Run a multilevel model to predict prose recall from age, including a random intercept and random slope. Be sure to include person-centered age (centered at age at the initial timepoint) and age at the initial timepoint as predictors. Note that age at the initial timepoint is already provided in the dataset (*ageT0*). 

```{r}
# code here
```


## Part c) [1.5 points]

Report and interpret the fixed effects of the model. 

**(Intercept):**

**Fixed effect of age at the initial timepoint:**

**Fixed effect of age:**


# Extra Credit

## EC 1 [0.5 point]

Using the ICC calculated in Question 1a, calculate the cluster effect. (*Note: each class has 100 students in it*)

```{r}
# code
```


## EC 2 [0.5 point]

Using the model run in Question 1b, with a random intercept and random slope - report and interpret the correlation between random effects.





