---
title: "Final Project"
format: pdf
execute:
  echo: false
---

```{r}
library(lavaan)
library(dplyr)
```
```{r}
dat <- read.csv("../data/clean_data2.csv")

mydata <- dplyr::filter(dat, redcap_event_name %in% c(
    "elegibilidade_arm_1", 
    "desfechos_arm_1", "followup_arm_1")
  ) |> 
  dplyr::mutate(
    redcap_event_name = factor(dplyr::case_when(
      redcap_event_name == "elegibilidade_arm_1" ~ 1,
      redcap_event_name == "desfechos_arm_1" ~ 2,
      redcap_event_name == "followup_arm_1" ~ 3
    )))
```


## Introduction

a.	Provide the theoretical rationale for your study. Include a brief literature review. Clearly specify your research questions/hypotheses.

## Methods

a.	Briefly describe your research design, target population, sampling/recruitment methods, and procedure.
b.	Specify the measures you are using and provide information on reliability and validity if possible.
c.	Specify your analysis plan and outline how each hypothesis will be tested. Write out the equations for the models.
d.	If there are missing data in your dataset, indicate how they will be treated.

Possible questions:

- Are initial levels of anxiety/depression/beliefs/flexibility correlated with the rate of change in insomnia/depression/anxiety/beliefs symptoms over time?

- Do individuals who start with high levels of X show a slower/steeper decline in insomnia severity?

https://gabriellajg.github.io/EPSY-579-R-Cookbook-for-SEM/lavaan-lab-16-latent-growth-models.html

https://longitudinalanalysis.com/mixture-latent-growth-models-r-a-step-by-step-guide/

https://tdjorgensen.github.io/SEM-in-Ed-compendium/ch27.html

Mixture Latent Growth Mixture Model introduces a categorical variable that accoutns for subgroup membership

## Results

a.	Clearly present and interpret your results. Use subheadings when necessary. 
b.	Include tables and/or figures in APA format. 

## Discussion

a.	Provide a summary of your research findings (e.g., are your hypotheses supported)?
b.	Discuss the implications of your findings and limitations of the study.

